---
title: "LVAD Database Comparison Analysis"
output: html_document
date: "2026-02-12"
---

```{r setup, include=FALSE}
library(dplyr)
library(ggplot2)
library(lubridate)
library(patchwork)
library(scales)
library(tidycmprsk)
library(survival)
library(broom)
library(knitr)
library(tidyr)
library(ggsurvfit)
```


``` {r load data}
#SRTR DATA!

JustFormHRDataLink_2025Q3 = read_sas("~/Desktop/R Project Data/HRStatJust2509DanielAhn/JustFormHRDataLink_2025Q3", NULL)

JustFormHR_2025Q3 = read_sas("~/Desktop/R Project Data/HRStatJust2509DanielAhn/JustFormHR_2025Q3.sas7bdat", NULL)

RiskStratDataHR_2025Q3 = read_sas("~/Desktop/R Project Data/HRStatJust2509DanielAhn/RiskStratDataHR_2025Q3.sas7bdat", NULL)

JustFormHRStat1_2025Q3 = read_sas("~/Desktop/R Project Data/HRStatJust2509DanielAhn/JustFormHRStat1_2025Q3.sas7bdat", NULL)

JustFormHRStat2_2025Q3 = read_sas("~/Desktop/R Project Data/HRStatJust2509DanielAhn/JustFormHRStat2_2025Q3.sas7bdat", NULL)

JustFormHRStat3_2025Q3 = read_sas("~/Desktop/R Project Data/HRStatJust2509DanielAhn/JustFormHRStat3_2025Q3.sas7bdat", NULL)

JustFormHRStat4_2025Q3 = read_sas("~/Desktop/R Project Data/HRStatJust2509DanielAhn/JustFormHRStat4_2025Q3.sas7bdat", NULL)

ThorSupportDevice_2025Q3 = read_sas("~/Desktop/R Project Data/HRStatJust2509DanielAhn/ThorSupportDevice_2025Q3.sas7bdat", NULL)

cand_thor2025q2 = read_sas("~/Desktop/R Project Data/HRStatJust2509DanielAhn/cand_thor2025q2.sas7bdat", NULL)

#STAR DATA

thoracic_data_2025q2 <- read_sas("~/Desktop/R Project Data/2025 Q2 STAR File/thoracic_data_2025q2.sas7bdat", NULL)

thoracic_stat1_2025q2 <- read_sas("~/Desktop/R Project Data/2025 Q2 STAR File/thoracic_stat1_2025q2.sas7bdat", NULL)

thoracic_stat2_2025q2 <- read_sas("~/Desktop/R Project Data/2025 Q2 STAR File/thoracic_stat2_2025q2.sas7bdat", NULL)

thoracic_stat3_2025q2 <- read_sas("~/Desktop/R Project Data/2025 Q2 STAR File/thoracic_stat3_2025q2.sas7bdat", NULL)

thoracic_stat4_2025q2 <- read_sas("~/Desktop/R Project Data/2025 Q2 STAR File/thoracic_stat4_2025q2.sas7bdat", NULL)

thoracic_mcs_device_2025q2 <- read_sas("~/Desktop/R Project Data/2025 Q2 STAR File/thoracic_mcs_device_2025q2.sas7bdat", NULL)



```


```{r link data sets in srtr}

just_form_hr_data_link = JustFormHRDataLink_2025Q3
just_form_hr = JustFormHR_2025Q3

risk_strat_data_hr = RiskStratDataHR_2025Q3 %>% 
  mutate(ChangeDt = floor_date(ChangeDate, unit = "day")) %>%
  select(-ChangeDate) %>% 
  full_join(just_form_hr_data_link %>%
              dplyr::select(-ChangeDate, -InitialFormJustId), by = "WlregAuditId") %>% 
  mutate(PX_ID = px_id) %>% 
  dplyr::select(-px_id) %>%
  
  # removes entries that have valid WlregAuditId but missing JustId
  filter(!is.na(JustId)) %>%
  
  # there are instances in which two just. forms were submitted on the same day for a
  # single PX_ID. The time step in this process is one day, so these are removed
  group_by(PX_ID, ChangeDt) %>% slice_max(JustId) %>% 
  ungroup() %>% 
  
  dplyr::select(-RiskStratId, -WlregAuditId, -ChangeUserId, -ChangeRequestorId) %>% 
  select(-(ends_with("St"))) %>% 
  select(-(starts_with("CandHist"))) %>%
  select(-(starts_with("SenData"))) %>%
  select(-(ends_with("Perf"))) %>%
  select(-(starts_with("CPT"))) %>%
  relocate(ChangeDt:PX_ID)


# right_join to filter out erroneous forms, as above
just_form_hr_data_link = just_form_hr_data_link %>%
  select(-ChangeDate) %>% 
  right_join(risk_strat_data_hr %>% 
              select(ChangeDt, JustId, PX_ID), by = "JustId")

# right_join to filter out erroneous forms, as above
just_form_hr = just_form_hr %>% 
  select(JustId, RequestedCandStatCd, RequestedCandStat_descrip, ExtensionNumber, Exception, ThorSupportDevicePrimary, ThorSupportDeviceSecondary, descrip, FormAddDt, FormEffectiveDt, FormExpirationDt, FormReceiptDt, FormStatus_descrip, ApplicationStatus_descrip, ChangeDate) %>%
  mutate(ChangeDt = floor_date(ChangeDate, unit = "day")) %>%
  mutate(listing_description = descrip) %>% 
  select(-descrip) %>% 
  right_join(just_form_hr_data_link %>% 
              select(-WlregAuditId, -InitialFormJustId), by = "JustId")


thor_support_device = ThorSupportDevice_2025Q3 %>%
  mutate(ChangeDt = as.Date(ChangeDate)) %>%
  left_join(just_form_hr %>% select(ThorSupportDevicePrimary, PX_ID),
            by = c("DeviceId" = "ThorSupportDevicePrimary")) %>%
  select(-DeviceId) %>%
  distinct(.keep_all = TRUE)

  # status_just_episode provides same info as stathist_thor, so not including it
  # # join status episode information by JustId
  # full_join(status_just_episode %>% select(JustId:EndDate), by = "JustId")

# thor_support_device seems to include many duplicates, include at your own risk

just_form_hr_stat1 = JustFormHRStat1_2025Q3 %>% 
  select(-ChangeDate) %>%
  left_join(just_form_hr_data_link %>% 
              select(JustId, PX_ID, ChangeDt), by = "JustId")

just_form_hr_stat2 = JustFormHRStat2_2025Q3 %>% 
  select(-ChangeDate) %>%
  left_join(just_form_hr_data_link %>% 
              select(JustId, PX_ID, ChangeDt), by = "JustId")

just_form_hr_stat3 = JustFormHRStat3_2025Q3 %>% 
  select(-ChangeDate) %>%
  left_join(just_form_hr_data_link %>% 
              select(JustId, PX_ID, ChangeDt), by = "JustId")

just_form_hr_stat4 = JustFormHRStat4_2025Q3 %>% 
  select(-ChangeDate) %>%
  left_join(just_form_hr_data_link %>% 
              select(JustId, PX_ID, ChangeDt), by = "JustId")

```

# SRTR LVAD Analysis (No Explant Logic)

```{r srtr_lvad}
# ============================================================================
# GET SRTR LVAD TIMELINE - NO EXPLANT LOGIC
# ============================================================================

# lvad_keywords <- c("heartmate", "hm", "lionheart", "arrow", "jarvik", "micromed", 
#                    "debakey", "novacor", "evaheart", "worldheart", "levacor", 
#                    "reliantheart", "hvad", "heartsaver")
# 
# srtr_lvad_implants <- thor_support_device %>%
#   filter(VadBrandId %in% c('202', '205', '206', '207', '208', '209', '210', '212', 
#                            '213', '214', '223', '224', '233', '236', '239', '240', 
#                            '312', '313', '314', '315', '316', '319', '322', '327', 
#                            '330', '333', '334') | 
#            (VadBrandId == 999 & grepl(paste(lvad_keywords, collapse = "|"), 
#                                       OtherSpecify, ignore.case = TRUE))) %>%
#   group_by(PX_ID) %>%
#   arrange(ImplantDt) %>%
#   slice(1) %>%
#   ungroup() %>%
#   dplyr::select(PX_ID, lvad_implant_date = ImplantDt)






# SRTR LVAD - simplified to match STAR coding
srtr_lvad_implants <- thor_support_device %>%
  filter(DeviceTypeId == 12 & VentricularSupportId == 1) %>%
  group_by(PX_ID) %>%
  arrange(ImplantDt) %>%
  slice(1) %>%
  ungroup() %>%
  dplyr::select(PX_ID, lvad_implant_date = ImplantDt)

cat("SRTR LVADs with simplified coding:", nrow(srtr_lvad_implants), "\n")

srtr_lvad_timeline <- srtr_lvad_implants

cat("Total SRTR patients with LVAD:", nrow(srtr_lvad_timeline), "\n")

# ============================================================================
# PREVALENCE ANALYSIS
# ============================================================================

dates_to_check <- c(
  mdy("10-18-2018"),
  seq(mdy("01-01-2019"), mdy("05-01-2025"), by = "2 months") - days(1)
)

get_srtr_lvad_prevalence_no_explant <- function(check_date) {
  candidates_on_date <- cand_thor2025q2 %>%
    filter(CAN_LISTING_DT <= check_date) %>%
    filter(is.na(CAN_REM_DT) | CAN_REM_DT > check_date) %>%
    filter(CAN_AGE_IN_MONTHS_AT_LISTING / 12 > 17) %>%
    dplyr::select(PX_ID)
  
  candidates_with_lvad <- candidates_on_date %>%
    left_join(srtr_lvad_timeline %>% dplyr::select(PX_ID, lvad_implant_date), by = "PX_ID") %>%
    mutate(has_lvad = !is.na(lvad_implant_date) & lvad_implant_date <= check_date)
  
  return(data.frame(
    date = check_date,
    total_waitlist = nrow(candidates_on_date),
    total_with_lvad = sum(candidates_with_lvad$has_lvad, na.rm = TRUE)
  ))
}

srtr_prevalence_list <- lapply(dates_to_check, get_srtr_lvad_prevalence_no_explant)
prevalence_results_df <- bind_rows(srtr_prevalence_list) %>%
  mutate(
    lvad_proportion = total_with_lvad / total_waitlist * 100,
    date_label = format(date, "%b %d, %Y"),
    date_factor = factor(date_label, levels = date_label)
  )

# ============================================================================
# FLOW/INCIDENCE ANALYSIS
# ============================================================================

period_starts <- seq(mdy("01-01-2019"), mdy("03-01-2025"), by = "2 months")
period_ends <- period_starts %m+% months(2) - days(1)

get_srtr_lvad_flow <- function(start_date, end_date) {
  candidates_in_period <- cand_thor2025q2 %>%
    filter(CAN_LISTING_DT <= end_date) %>%
    filter(is.na(CAN_REM_DT) | CAN_REM_DT >= start_date) %>%
    filter(CAN_AGE_IN_MONTHS_AT_LISTING / 12 > 17) %>%
    dplyr::select(PX_ID, CAN_LISTING_DT, CAN_REM_DT)
  
  candidates_with_lvad <- candidates_in_period %>%
    left_join(srtr_lvad_timeline, by = "PX_ID")
  
  new_listings_with_lvad <- candidates_with_lvad %>%
    filter(CAN_LISTING_DT >= start_date & CAN_LISTING_DT <= end_date) %>%
    filter(!is.na(lvad_implant_date) & lvad_implant_date <= CAN_LISTING_DT) %>%
    nrow()
  
  lvad_implants_on_waitlist <- candidates_with_lvad %>%
    filter(!is.na(lvad_implant_date)) %>%
    filter(lvad_implant_date >= start_date & lvad_implant_date <= end_date) %>%
    filter(lvad_implant_date > CAN_LISTING_DT) %>%
    nrow()
  
  total_new_listings <- candidates_with_lvad %>%
    filter(CAN_LISTING_DT >= start_date & CAN_LISTING_DT <= end_date) %>%
    nrow()
  
  return(data.frame(
    period_start = start_date,
    period_end = end_date,
    new_listings_with_lvad = new_listings_with_lvad,
    lvad_implants_on_waitlist = lvad_implants_on_waitlist,
    total_new_listings = total_new_listings
  ))
}

flow_results_list <- lapply(1:length(period_starts), function(i) {
  get_srtr_lvad_flow(period_starts[i], period_ends[i])
})

flow_results_df <- bind_rows(flow_results_list) %>%
  mutate(
    period_label = paste(format(period_start, "%b"), format(period_end, "%b %Y"), sep = "-"),
    period_factor = factor(period_label, levels = period_label),
    pct_listed_with_lvad = (new_listings_with_lvad / total_new_listings) * 100
  )

# ============================================================================
# VISUALIZATIONS
# ============================================================================

# Prevalence plots
plot_prev_a <- ggplot(prevalence_results_df, aes(x = date_factor, group = 1)) +
  geom_line(aes(y = total_waitlist, color = "Total Waitlist"), linewidth = 1) +
  geom_point(aes(y = total_waitlist, color = "Total Waitlist"), size = 3) +
  geom_line(aes(y = total_with_lvad, color = "LVAD Support"), linewidth = 1) +
  geom_point(aes(y = total_with_lvad, color = "LVAD Support"), size = 3) +
  scale_color_manual(values = c("Total Waitlist" = "steelblue", "LVAD Support" = "coral")) +
  scale_y_continuous(labels = comma, limits = c(0, NA)) +
  labs(
    title = "SRTR: Waitlist and LVAD-Supported Candidates Over Time",
    x = "Date",
    y = "Number of Candidates",
    color = NULL
  ) +
  theme_classic(base_size = 15) +
  theme(
    legend.position = "bottom",
    axis.text.x = element_text(angle = 45, hjust = 1, size = 15),
    axis.text.y = element_text(size = 15),
    axis.title = element_text(size = 15),
    plot.title = element_text(size = 15),
    legend.text = element_text(size = 15)
  )

plot_prev_b <- ggplot(prevalence_results_df, aes(x = date_factor, y = lvad_proportion)) +
  geom_col(fill = "coral", alpha = 0.7) +
  scale_y_continuous(limits = c(0, NA)) +
  labs(
    title = "SRTR: Proportion of Waitlist with LVAD Support",
    x = "Date",
    y = "Percentage with LVAD (%)"
  ) +
  theme_classic(base_size = 15) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 15),
    axis.text.y = element_text(size = 15),
    axis.title = element_text(size = 15),
    plot.title = element_text(size = 15)
  )

combined_prevalence_plot <- plot_prev_a / plot_prev_b

# Flow plots
flow_plot_a <- ggplot(flow_results_df, aes(x = period_factor, group = 1)) +
  geom_line(aes(y = new_listings_with_lvad, color = "Listed with LVAD"), linewidth = 1) +
  geom_point(aes(y = new_listings_with_lvad, color = "Listed with LVAD"), size = 3) +
  geom_line(aes(y = lvad_implants_on_waitlist, color = "LVAD Implanted While Listed"), linewidth = 1) +
  geom_point(aes(y = lvad_implants_on_waitlist, color = "LVAD Implanted While Listed"), size = 3) +
  scale_color_manual(values = c("Listed with LVAD" = "darkgreen", 
                                "LVAD Implanted While Listed" = "purple")) +
  scale_y_continuous(labels = comma, limits = c(0, NA)) +
  labs(
    title = "SRTR: LVAD Flow Dynamics",
    x = "Time Period",
    y = "Number of Candidates",
    color = NULL
  ) +
  theme_classic(base_size = 15) +
  theme(
    legend.position = "bottom",
    axis.text.x = element_text(angle = 45, hjust = 1, size = 15),
    axis.text.y = element_text(size = 15),
    axis.title = element_text(size = 15),
    plot.title = element_text(size = 15),
    legend.text = element_text(size = 12)
  )

# Display plots
print(combined_prevalence_plot)
print(flow_plot_a)

# Print summaries
cat("\n=== SRTR PREVALENCE SUMMARY ===\n")
print(prevalence_results_df %>%
        dplyr::select(date_label, total_waitlist, total_with_lvad, lvad_proportion) %>%
        mutate(lvad_proportion = round(lvad_proportion, 2)))

cat("\n=== SRTR FLOW SUMMARY ===\n")
print(flow_results_df %>%
        dplyr::select(period_label, new_listings_with_lvad, lvad_implants_on_waitlist, 
               total_new_listings, pct_listed_with_lvad) %>%
        mutate(pct_listed_with_lvad = round(pct_listed_with_lvad, 2)))
```

# STAR MCS File Analysis (No Explant Logic)

```{r star_mcs}
# ============================================================================
# BUILD STAR MCS FILE COHORT
# ============================================================================

lvads_from_mcs_file_optn <- thoracic_mcs_device_2025q2 %>%
  filter(VENTRICULAR_SUPPORT == "Left") %>%
  filter(DEVICE_TYPE == "Dischargeable VAD") %>%
  arrange(WL_ID_CODE, IMPLANT_DATE, desc(EXPLANT_DATE)) %>%
  distinct(WL_ID_CODE, IMPLANT_DATE, .keep_all = TRUE)

star_lvad_mcs <- lvads_from_mcs_file_optn %>%
  group_by(WL_ID_CODE) %>%
  arrange(IMPLANT_DATE) %>%
  summarize(
    first_lvad_implant = first(IMPLANT_DATE),
    .groups = "drop"
  ) %>%
  mutate(first_lvad_implant = as.Date(first_lvad_implant, origin = "1970-01-01"))

cat("STAR MCS file: Total patients with LVAD:", nrow(star_lvad_mcs), "\n")

# ============================================================================
# PREVALENCE ANALYSIS
# ============================================================================

get_star_mcs_lvad_prevalence <- function(check_date) {
  candidates_on_date <- thoracic_data_2025q2 %>%
    filter(INIT_DATE <= check_date) %>%
    filter(is.na(END_DATE) | END_DATE > check_date) %>%
    filter(INIT_AGE > 17) %>%
    dplyr::select(WL_ID_CODE)
  
  candidates_with_lvad <- candidates_on_date %>%
    left_join(star_lvad_mcs %>% dplyr::select(WL_ID_CODE, first_lvad_implant), 
              by = "WL_ID_CODE") %>%
    mutate(has_lvad = !is.na(first_lvad_implant) & first_lvad_implant <= check_date)
  
  return(data.frame(
    date = check_date,
    total_waitlist = nrow(candidates_on_date),
    total_with_lvad = sum(candidates_with_lvad$has_lvad, na.rm = TRUE)
  ))
}

star_prevalence_list <- lapply(dates_to_check, get_star_mcs_lvad_prevalence)
star_prevalence_df <- bind_rows(star_prevalence_list) %>%
  mutate(
    lvad_proportion = total_with_lvad / total_waitlist * 100,
    date_label = format(date, "%b %d, %Y"),
    date_factor = factor(date_label, levels = date_label)
  )

# ============================================================================
# FLOW/INCIDENCE ANALYSIS
# ============================================================================

get_star_lvad_flow <- function(start_date, end_date) {
  candidates_in_period <- thoracic_data_2025q2 %>%
    filter(INIT_DATE <= end_date) %>%
    filter(is.na(END_DATE) | END_DATE > start_date) %>%
    filter(INIT_AGE > 17) %>%
    dplyr::select(WL_ID_CODE, INIT_DATE, END_DATE)
  
  candidates_with_lvad <- candidates_in_period %>%
    left_join(star_lvad_mcs, by = "WL_ID_CODE")
  
  new_listings_with_lvad <- candidates_with_lvad %>%
    filter(INIT_DATE >= start_date & INIT_DATE <= end_date) %>%
    filter(!is.na(first_lvad_implant) & first_lvad_implant <= INIT_DATE) %>%
    nrow()
  
  lvad_implants_on_waitlist <- candidates_with_lvad %>%
    filter(!is.na(first_lvad_implant)) %>%
    filter(first_lvad_implant >= start_date & first_lvad_implant <= end_date) %>%
    filter(first_lvad_implant > INIT_DATE) %>%
    nrow()
  
  total_new_listings <- candidates_with_lvad %>%
    filter(INIT_DATE >= start_date & INIT_DATE <= end_date) %>%
    nrow()
  
  return(data.frame(
    period_start = start_date,
    period_end = end_date,
    new_listings_with_lvad = new_listings_with_lvad,
    lvad_implants_on_waitlist = lvad_implants_on_waitlist,
    total_new_listings = total_new_listings
  ))
}

star_flow_list <- lapply(1:length(period_starts), function(i) {
  get_star_lvad_flow(period_starts[i], period_ends[i])
})

star_flow_df <- bind_rows(star_flow_list) %>%
  mutate(
    period_label = paste(format(period_start, "%b"), format(period_end, "%b %Y"), sep = "-"),
    period_factor = factor(period_label, levels = period_label),
    pct_listed_with_lvad = (new_listings_with_lvad / total_new_listings) * 100
  )

# ============================================================================
# VISUALIZATIONS
# ============================================================================

star_plot_prev_a <- ggplot(star_prevalence_df, aes(x = date_factor, group = 1)) +
  geom_line(aes(y = total_waitlist, color = "Total Waitlist"), linewidth = 1) +
  geom_point(aes(y = total_waitlist, color = "Total Waitlist"), size = 3) +
  geom_line(aes(y = total_with_lvad, color = "LVAD Support"), linewidth = 1) +
  geom_point(aes(y = total_with_lvad, color = "LVAD Support"), size = 3) +
  scale_color_manual(values = c("Total Waitlist" = "steelblue", "LVAD Support" = "coral")) +
  scale_y_continuous(labels = comma, limits = c(0, NA)) +
  labs(
    title = "STAR: Waitlist and LVAD-Supported Candidates Over Time",
    x = "Date",
    y = "Number of Candidates",
    color = NULL
  ) +
  theme_classic(base_size = 15) +
  theme(
    legend.position = "bottom",
    axis.text.x = element_text(angle = 45, hjust = 1, size = 15),
    axis.text.y = element_text(size = 15),
    axis.title = element_text(size = 15),
    plot.title = element_text(size = 15),
    legend.text = element_text(size = 15)
  )

star_plot_prev_b <- ggplot(star_prevalence_df, aes(x = date_factor, y = lvad_proportion)) +
  geom_col(fill = "coral", alpha = 0.7) +
  scale_y_continuous(limits = c(0, NA)) +
  labs(
    title = "STAR: Proportion of Waitlist with LVAD Support",
    x = "Date",
    y = "Percentage with LVAD (%)"
  ) +
  theme_classic(base_size = 15) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 15),
    axis.text.y = element_text(size = 15),
    axis.title = element_text(size = 15),
    plot.title = element_text(size = 15)
  )

combined_star_prevalence_plot <- star_plot_prev_a / star_plot_prev_b

star_flow_plot_a <- ggplot(star_flow_df, aes(x = period_factor, group = 1)) +
  geom_line(aes(y = new_listings_with_lvad, color = "Listed with LVAD"), linewidth = 1) +
  geom_point(aes(y = new_listings_with_lvad, color = "Listed with LVAD"), size = 3) +
  geom_line(aes(y = lvad_implants_on_waitlist, color = "LVAD Implanted While Listed"), linewidth = 1) +
  geom_point(aes(y = lvad_implants_on_waitlist, color = "LVAD Implanted While Listed"), size = 3) +
  scale_color_manual(values = c("Listed with LVAD" = "darkgreen", 
                                "LVAD Implanted While Listed" = "purple")) +
  scale_y_continuous(labels = comma, limits = c(0, NA)) +
  labs(
    title = "STAR: LVAD Flow Dynamics",
    x = "Time Period",
    y = "Number of Candidates",
    color = NULL
  ) +
  theme_classic(base_size = 15) +
  theme(
    legend.position = "bottom",
    axis.text.x = element_text(angle = 45, hjust = 1, size = 15),
    axis.text.y = element_text(size = 15),
    axis.title = element_text(size = 15),
    plot.title = element_text(size = 15),
    legend.text = element_text(size = 12)
  )

print(combined_star_prevalence_plot)
print(star_flow_plot_a)

cat("\n=== STAR PREVALENCE SUMMARY ===\n")
print(star_prevalence_df %>%
        dplyr::select(date_label, total_waitlist, total_with_lvad, lvad_proportion) %>%
        mutate(lvad_proportion = round(lvad_proportion, 2)))

cat("\n=== STAR FLOW SUMMARY ===\n")
print(star_flow_df %>%
        dplyr::select(period_label, new_listings_with_lvad, lvad_implants_on_waitlist,
               total_new_listings, pct_listed_with_lvad) %>%
        mutate(pct_listed_with_lvad = round(pct_listed_with_lvad, 2)))
```

# SRTR vs STAR MCS Direct Comparison

```{r srtr_vs_star_comparison}
# ============================================================================
# COMBINE PREVALENCE DATA
# ============================================================================

combined_prevalence <- bind_rows(
  prevalence_results_df %>%
    dplyr::select(date, date_label, date_factor, total_waitlist, total_with_lvad, lvad_proportion) %>%
    mutate(data_source = "SRTR (MCS File)"),
  star_prevalence_df %>%
    dplyr::select(date, date_label, date_factor, total_waitlist, total_with_lvad, lvad_proportion) %>%
    mutate(data_source = "STAR (MCS File)")
) %>%
  mutate(data_source = factor(data_source, levels = c("SRTR (MCS File)", "STAR (MCS File)")))

# ============================================================================
# PREVALENCE COMPARISON PLOTS
# ============================================================================

prev_comparison_lvad <- ggplot(combined_prevalence, aes(x = date_factor, group = data_source)) +
  geom_line(aes(y = total_with_lvad, color = data_source, linetype = data_source), linewidth = 1.2) +
  geom_point(aes(y = total_with_lvad, color = data_source, shape = data_source), size = 2.5) +
  scale_color_manual(values = c("SRTR (MCS File)" = "coral", "STAR (MCS File)" = "steelblue")) +
  scale_linetype_manual(values = c("SRTR (MCS File)" = "solid", "STAR (MCS File)" = "dashed")) +
  scale_shape_manual(values = c("SRTR (MCS File)" = 16, "STAR (MCS File)" = 17)) +
  scale_y_continuous(labels = comma, limits = c(0, NA)) +
  labs(
    x = "",
    y = "Number of Candidates\nwith LVADs",
    color = "Data Source",
    linetype = "Data Source",
    shape = "Data Source"
  ) +
  theme_classic(base_size = 15) +
  theme(
    legend.position = "bottom",
    axis.text.x = element_text(angle = 45, hjust = 1, size = 15),
    axis.text.y = element_text(size = 15),
    axis.title = element_text(size = 15),
    legend.text = element_text(size = 15)
  )

prev_comparison_proportion <- ggplot(combined_prevalence, aes(x = date_factor, y = lvad_proportion, fill = data_source)) +
  geom_col(position = "dodge", alpha = 0.7) +
  scale_fill_manual(values = c("SRTR (MCS File)" = "coral", "STAR (MCS File)" = "steelblue")) +
  scale_y_continuous(limits = c(0, NA)) +
  labs(
    x = "Date",
    y = "% of Waitlist\nwith LVADs",
    fill = "Data Source"
  ) +
  theme_classic(base_size = 15) +
  theme(
    legend.position = "bottom",
    axis.text.x = element_text(angle = 45, hjust = 1, size = 15),
    axis.text.y = element_text(size = 15),
    axis.title = element_text(size = 15),
    legend.text = element_text(size = 15)
  )

combined_prev_comparison_plot <- (prev_comparison_lvad + ggtitle("A")) / 
  (prev_comparison_proportion + ggtitle("B")) +
  plot_layout(guides = "collect") &
  theme(
    legend.position = "bottom",
    plot.title = element_text(size = 18, face = "bold", hjust = 0)
  )

print(combined_prev_comparison_plot)

# ============================================================================
# DISCREPANCY STATISTICS
# ============================================================================

prevalence_discrepancy <- combined_prevalence %>%
  dplyr::select(date, date_label, data_source, total_with_lvad, lvad_proportion) %>%
  tidyr::pivot_wider(
    names_from = data_source,
    values_from = c(total_with_lvad, lvad_proportion)
  ) %>%
  mutate(
    absolute_missing = `total_with_lvad_SRTR` - `total_with_lvad_OPTN STAR`,
    pct_missing = (absolute_missing / `total_with_lvad_SRTR`) * 100,
    proportion_gap = `lvad_proportion_SRTR` - `lvad_proportion_OPTN STAR`
  )

print(prevalence_discrepancy)

cat("\n=== PREVALENCE UNDERREPORTING IN STAR ===\n")
cat("\nOverall statistics:\n")
cat("Mean LVAD patients missing in STAR:", round(mean(prevalence_discrepancy$absolute_missing, na.rm = TRUE), 0), "\n")
cat("Mean % underreported:", round(mean(prevalence_discrepancy$pct_missing, na.rm = TRUE), 1), "%\n")
cat("Median % underreported:", round(median(prevalence_discrepancy$pct_missing, na.rm = TRUE), 1), "%\n")
```

# Flow Comparison: SRTR vs STAR MCS

```{r flow_comparison}
# ============================================================================
# COMBINE FLOW DATA
# ============================================================================

srtr_flow_comparison <- flow_results_df %>%
  dplyr::select(period_label, period_factor, new_listings_with_lvad, lvad_implants_on_waitlist) %>%
  mutate(data_source = "SRTR (MCS File)")

star_flow_comparison <- star_flow_df %>%
  dplyr::select(period_label, period_factor, new_listings_with_lvad, lvad_implants_on_waitlist) %>%
  mutate(data_source = "STAR (MCS File)")

combined_flow <- bind_rows(srtr_flow_comparison, star_flow_comparison) %>%
  mutate(data_source = factor(data_source, levels = c("SRTR (MCS File)", "STAR (MCS File)")))

# ============================================================================
# FLOW COMPARISON PLOT
# ============================================================================

combined_flow_plot <- ggplot(combined_flow, aes(x = period_factor, group = interaction(data_source))) +
  geom_line(aes(y = new_listings_with_lvad, color = data_source, linetype = "Listed with LVAD"), 
            linewidth = 1.2) +
  geom_point(aes(y = new_listings_with_lvad, color = data_source, shape = data_source), 
             size = 2.5) +
  geom_line(aes(y = lvad_implants_on_waitlist, color = data_source, linetype = "LVAD Implanted While Listed"), 
            linewidth = 1.2) +
  geom_point(aes(y = lvad_implants_on_waitlist, color = data_source, shape = data_source), 
             size = 2.5) +
  scale_color_manual(values = c("SRTR (MCS File)" = "coral", "STAR (MCS File)" = "steelblue"),
                     name = "Data Source") +
  scale_linetype_manual(values = c("Listed with LVAD" = "solid", 
                                   "LVAD Implanted While Listed" = "dotted"),
                        name = "Event Type") +
  scale_shape_manual(values = c("SRTR (MCS File)" = 16, "STAR (MCS File)" = 17),
                     name = "Data Source") +
  scale_y_continuous(labels = comma, limits = c(0, NA)) +
  labs(
    x = "Time Period",
    y = "Number of Candidates"
  ) +
  theme_classic(base_size = 15) +
  theme(
    legend.position = "bottom",
    legend.box = "vertical",
    axis.text.x = element_text(angle = 45, hjust = 1, size = 15),
    axis.text.y = element_text(size = 15),
    axis.title = element_text(size = 15),
    legend.text = element_text(size = 15)
  )

print(combined_flow_plot)

# ============================================================================
# FLOW DISCREPANCY STATISTICS
# ============================================================================

flow_discrepancy <- combined_flow %>%
  dplyr::select(period_label, data_source, new_listings_with_lvad, lvad_implants_on_waitlist) %>%
  pivot_wider(
    names_from = data_source,
    values_from = c(new_listings_with_lvad, lvad_implants_on_waitlist)
  ) %>%
  mutate(
    listed_missing = `new_listings_with_lvad_SRTR` - `new_listings_with_lvad_OPTN STAR`,
    listed_pct_missing = (listed_missing / `new_listings_with_lvad_SRTR`) * 100,
    implanted_diff = `lvad_implants_on_waitlist_SRTR` - `lvad_implants_on_waitlist_OPTN STAR`
  )

cat("\n=== FLOW UNDERREPORTING IN STAR ===\n")
cat("\nListed with LVAD:\n")
cat("Mean missing in STAR:", round(mean(flow_discrepancy$listed_missing, na.rm = TRUE), 0), "\n")
cat("Mean % underreported:", round(mean(flow_discrepancy$listed_pct_missing, na.rm = TRUE), 1), "%\n")

cat("\nLVAD Implanted While Listed:\n")
cat("Mean difference:", round(mean(flow_discrepancy$implanted_diff, na.rm = TRUE), 0), "\n")
```


# Status 5/6 Validation Analysis

```{r status56_validation}
# ============================================================================
# VALIDATE STATUS 5/6 AS EXPLANT PROXY
# ============================================================================

# Build MCS cohort with explant information
mcs_file_cohort <- star_lvads_from_justifications %>%
  filter(!is.na(INPLACE) | !is.na(EXPLANT_DATE)) %>%
  distinct(WL_ID_CODE, .keep_all = TRUE) %>%
  mutate(
    lvad_implant = as.Date(PRIM_IMPLANTDT, origin = "1970-01-01"),
    explant_date = as.Date(EXPLANT_DATE, origin = "1970-01-01"),
    has_explant = !is.na(EXPLANT_DATE) & MCS_REM_REASON != "Transplanted"
  ) %>%
  dplyr::select(WL_ID_CODE, lvad_implant, explant_date, has_explant, MCS_REM_REASON)

cat("MCS file cohort (delisted patients):", nrow(mcs_file_cohort), "\n")
cat("With documented non-transplant explants:", sum(mcs_file_cohort$has_explant), "\n\n")

# ============================================================================
# SENSITIVITY: Of true explants, how many went to Status 5/6?
# ============================================================================

true_explants <- mcs_file_cohort %>%
  filter(has_explant) %>%
  left_join(
    thoracic_wlhistory_data_2025q2 %>%
      dplyr::select(WL_ID_CODE, UNOS_CAND_STAT_CD, CHG_DATE),
    by = "WL_ID_CODE"
  ) %>%
  filter(CHG_DATE >= lvad_implant & CHG_DATE <= explant_date) %>%
  mutate(is_status_56 = UNOS_CAND_STAT_CD %in% c(1150, 1160, 2150, 2160)) %>%
  group_by(WL_ID_CODE, lvad_implant, explant_date, MCS_REM_REASON) %>%
  summarize(went_to_status_56 = any(is_status_56), .groups = "drop")

cat("=== SENSITIVITY ANALYSIS ===\n")
cat("Documented explants:", nrow(true_explants), "\n")
cat("Went to Status 5/6 before explant:", sum(true_explants$went_to_status_56), "\n")
cat("Never went to Status 5/6:", sum(!true_explants$went_to_status_56), "\n")
cat("Sensitivity:", round(sum(true_explants$went_to_status_56) / nrow(true_explants) * 100, 1), "%\n\n")

# Breakdown by removal reason
cat("By removal reason:\n")
print(true_explants %>%
  group_by(MCS_REM_REASON) %>%
  summarize(
    total = n(),
    went_to_56 = sum(went_to_status_56),
    pct_56 = round(sum(went_to_status_56) / n() * 100, 1)
  ) %>%
  arrange(desc(total)))

# ============================================================================
# POSITIVE PREDICTIVE VALUE: Of Status 5/6 transitions, how many had explants?
# ============================================================================

status56_in_cohort <- mcs_file_cohort %>%
  left_join(
    thoracic_wlhistory_data_2025q2 %>%
      dplyr::select(WL_ID_CODE, UNOS_CAND_STAT_CD, CHG_DATE),
    by = "WL_ID_CODE"
  ) %>%
  filter(CHG_DATE >= lvad_implant) %>%
  mutate(is_status_56 = UNOS_CAND_STAT_CD %in% c(1150, 1160, 2150, 2160)) %>%
  group_by(WL_ID_CODE, lvad_implant, explant_date, has_explant) %>%
  summarize(went_to_status_56 = any(is_status_56), .groups = "drop") %>%
  filter(went_to_status_56)

cat("\n=== POSITIVE PREDICTIVE VALUE ===\n")
cat("Patients who went to Status 5/6:", nrow(status56_in_cohort), "\n")
cat("Had documented explant:", sum(status56_in_cohort$has_explant), "\n")
cat("Never had explant:", sum(!status56_in_cohort$has_explant), "\n")
cat("PPV (% with actual explant):", 
    round(sum(status56_in_cohort$has_explant) / nrow(status56_in_cohort) * 100, 1), "%\n\n")

# ============================================================================
# SPECIFICITY CALCULATION
# ============================================================================

without_explant <- sum(!mcs_file_cohort$has_explant)
went_to_56_no_explant <- sum(!status56_in_cohort$has_explant)
never_56_no_explant <- without_explant - went_to_56_no_explant

cat("=== SPECIFICITY ===\n")
cat("Patients without explant:", without_explant, "\n")
cat("Went to Status 5/6 (false positive):", went_to_56_no_explant, "\n")
cat("Never went to Status 5/6 (true negative):", never_56_no_explant, "\n")
cat("Specificity:", round(never_56_no_explant / without_explant * 100, 1), "%\n\n")

cat("=== SUMMARY ===\n")
cat("Status 5/6 as explant proxy:\n")
cat("  Sensitivity: 29.9% (misses 70% of true explants)\n")
cat("  Specificity: 75.3% (24.7% false positive rate)\n")
cat("  Positive Predictive Value: 2.3%\n")
cat("  Interpretation: Status 5/6 is an unreliable explant indicator\n")
```

# Outcomes Analysis: SRTR vs STAR MCS (3-Year Follow-up)

```{r outcomes_srtr_vs_star}
# ============================================================================
# BUILD OUTCOMES COHORTS - PROPERLY CENSORED AT 3 YEARS
# ============================================================================

# SRTR cohort
srtr_lvad_outcomes <- srtr_lvad_timeline %>%
  filter(lvad_implant_date >= mdy("10-18-2018") &
         lvad_implant_date <= mdy("12-31-2024")) %>%
  left_join(cand_thor2025q2 %>%
              dplyr::select(PX_ID, CAN_LISTING_DT, CAN_REM_DT, CAN_REM_CD,
                           CAN_AGE_IN_MONTHS_AT_LISTING),
            by = "PX_ID") %>%
  filter(CAN_AGE_IN_MONTHS_AT_LISTING / 12 > 17) %>%
  mutate(
    removal_or_censor_date = if_else(is.na(CAN_REM_DT),
                                     as.Date("2025-06-30"),
                                     as.Date(CAN_REM_DT)),
    end_date = removal_or_censor_date,
    followup_time_original = as.numeric(difftime(end_date, lvad_implant_date, units = "days")) / 365.25,
    followup_time = pmin(followup_time_original, 3),  # Cap at 3 years
    event = case_when(
      followup_time_original > 3 ~ 0,  # Censor if event occurred after 3 years
      CAN_REM_CD %in% c(8, 13) ~ 1,
      CAN_REM_CD == 4 ~ 2,
      TRUE ~ 0
    ),
    event_factor = factor(event, levels = c(0, 1, 2),
                          labels = c("Censored", "Death/Deterioration", "Transplant")),
    data_source = "SRTR (MCS File)"
  ) %>%
  filter(followup_time > 0)

# STAR MCS cohort
star_lvad_outcomes <- star_lvad_mcs %>%
  filter(first_lvad_implant >= mdy("10-18-2018") &
         first_lvad_implant <= mdy("12-31-2024")) %>%
  left_join(thoracic_data_2025q2 %>%
              dplyr::select(WL_ID_CODE, INIT_DATE, END_DATE, REM_CD, INIT_AGE),
            by = "WL_ID_CODE") %>%
  filter(INIT_AGE > 17) %>%
  mutate(
    removal_or_censor_date = if_else(is.na(END_DATE),
                                     as.Date("2025-06-30"),
                                     as.Date(END_DATE)),
    end_date = removal_or_censor_date,
    followup_time_original = as.numeric(difftime(end_date, first_lvad_implant, units = "days")) / 365.25,
    followup_time = pmin(followup_time_original, 3),  # Cap at 3 years
    event = case_when(
      followup_time_original > 3 ~ 0,  # Censor if event occurred after 3 years
      REM_CD %in% c(8, 13) ~ 1,
      REM_CD == 4 ~ 2,
      TRUE ~ 0
    ),
    event_factor = factor(event, levels = c(0, 1, 2),
                          labels = c("Censored", "Death/Deterioration", "Transplant")),
    data_source = "STAR (MCS File)"
  ) %>%
  filter(followup_time > 0)

cat("SRTR outcomes cohort:", nrow(srtr_lvad_outcomes), "patients\n")
cat("  Deaths/Deterioration:", sum(srtr_lvad_outcomes$event == 1), "\n")
cat("  Transplants:", sum(srtr_lvad_outcomes$event == 2), "\n")
cat("  Censored:", sum(srtr_lvad_outcomes$event == 0), "\n\n")

cat("STAR outcomes cohort:", nrow(star_lvad_outcomes), "patients\n")
cat("  Deaths/Deterioration:", sum(star_lvad_outcomes$event == 1), "\n")
cat("  Transplants:", sum(star_lvad_outcomes$event == 2), "\n")
cat("  Censored:", sum(star_lvad_outcomes$event == 0), "\n")

# ============================================================================
# COMBINE AND ANALYZE
# ============================================================================

combined_outcomes <- bind_rows(
  srtr_lvad_outcomes %>% dplyr::select(followup_time, event_factor, data_source),
  star_lvad_outcomes %>% dplyr::select(followup_time, event_factor, data_source)
) %>%
  mutate(data_source = factor(data_source, levels = c("SRTR (MCS File)", "STAR (MCS File)")))

# Cumulative incidence
cif_fit_2db <- tidycmprsk::cuminc(Surv(followup_time, event_factor) ~ data_source,
                                   data = combined_outcomes)



# Fine-Gray tests
fg_transplant_2db <- tidycmprsk::crr(Surv(followup_time, event_factor) ~ data_source,
                                      data = combined_outcomes,
                                      failcode = "Transplant")
fg_death_2db <- tidycmprsk::crr(Surv(followup_time, event_factor) ~ data_source,
                                 data = combined_outcomes,
                                 failcode = "Death/Deterioration")

p_transplant_2db <- tidy(fg_transplant_2db)$p.value[1]
p_death_2db <- tidy(fg_death_2db)$p.value[1]

cat("\n=== FINE-GRAY TESTS: SRTR vs STAR ===\n")
cat("Transplant p:", format.pval(p_transplant_2db, digits = 3), "\n")
cat("Death/Deterioration p:", format.pval(p_death_2db, digits = 3), "\n")

# ============================================================================
# PLOT
# ============================================================================

fmt_p <- function(p) {
  if (p < 0.001) "p<0.001" else paste0("p=", format.pval(p, digits = 3))
}

cif_plot_2db <- cif_fit_2db %>%
  ggcuminc(
    outcome = c("Death/Deterioration", "Transplant"),
    linewidth = 1.2
  ) +
  add_confidence_interval(alpha = 0.2) +
  add_risktable(
    size = 5,
    theme = theme_risktable_default(
      plot.title.size = 14,
      axis.text.y.size = 14
    )
  ) +
  scale_x_continuous(breaks = 0:3, limits = c(0, 3)) +
  scale_y_continuous(
    limits = c(0, 1),
    labels = scales::percent,
    expand = c(0.01, 0)
  ) +
  xlab("Years from LVAD Implantation") +
  ylab("Cumulative Incidence") +
  scale_color_manual(
    values = c("SRTR (MCS File)" = "coral", "STAR (MCS File)" = "steelblue"),
    name = "Data Source"
  ) +
  scale_fill_manual(
    values = c("SRTR (MCS File)" = "coral", "STAR (MCS File)" = "steelblue"),
    name = "Data Source"
  ) +
  scale_linetype_discrete(name = "Outcome") +
  guides(
    color = guide_legend(order = 1, nrow = 1),
    fill = guide_legend(order = 1, nrow = 1),
    linetype = guide_legend(order = 2, nrow = 1)
  ) +
  # annotate("text", x = 2.8, y = 0.95,
  #          label = paste0("Fine-Gray Transplant: ", fmt_p(p_transplant_2db)),
  #          hjust = 1, size = 5, fontface = "bold") +
  # annotate("text", x = 2.8, y = 0.90,
  #          label = paste0("Fine-Gray Death/Deterioration: ", fmt_p(p_death_2db)),
  #          hjust = 1, size = 5, fontface = "bold") +
  theme_classic() +
  theme(
    legend.position = "bottom",
    legend.box = "vertical",
    legend.title = element_text(size = 15, face = "bold"),
    axis.text = element_text(size = 15),
    axis.title = element_text(size = 15),
    legend.text = element_text(size = 15)
  )

print(cif_plot_2db)


# ============================================================================
# EXTRACT CUMULATIVE INCIDENCE AT 3 YEARS
# ============================================================================

cif_summary_3yr <- cif_fit_2db %>%
  tidy() %>%
  filter(time <= 3.5) %>%  # Look within reasonable range of 3 years
  group_by(outcome, strata) %>%
  filter(abs(time - 3) == min(abs(time - 3))) %>%  # Find closest to 3 years
  ungroup() %>%
  mutate(
    time_year = round(time, 2),
    estimate_pct = round(estimate * 100, 1),
    conf_low_pct = round(conf.low * 100, 1),
    conf_high_pct = round(conf.high * 100, 1)
  ) %>%
  select(outcome, strata, time_year, estimate_pct, conf_low_pct, conf_high_pct)

cat("\n=== CUMULATIVE INCIDENCE AT 3 YEARS ===\n")
print(cif_summary_3yr)

# Alternatively, for a nicer formatted table:
cif_summary_3yr_wide <- cif_summary_3yr %>%
  mutate(
    ci = paste0(estimate_pct, "% (", conf_low_pct, "-", conf_high_pct, "%)")
  ) %>%
  select(outcome, strata, ci) %>%
  tidyr::pivot_wider(names_from = strata, values_from = ci)

cat("\n=== FORMATTED 3-YEAR CUMULATIVE INCIDENCE ===\n")
print(cif_summary_3yr_wide)





# Calculate median follow-up time

# SRTR cohort
srtr_median_followup <- median(srtr_lvad_outcomes$followup_time_original, na.rm = TRUE)
srtr_iqr_followup <- quantile(srtr_lvad_outcomes$followup_time_original, 
                               probs = c(0.25, 0.75), na.rm = TRUE)

cat("SRTR Median Follow-up:", round(srtr_median_followup, 2), "years\n")
cat("SRTR IQR:", round(srtr_iqr_followup[1], 2), "-", 
    round(srtr_iqr_followup[2], 2), "years\n\n")

# STAR cohort
star_median_followup <- median(star_lvad_outcomes$followup_time_original, na.rm = TRUE)
star_iqr_followup <- quantile(star_lvad_outcomes$followup_time_original, 
                               probs = c(0.25, 0.75), na.rm = TRUE)

cat("STAR Median Follow-up:", round(star_median_followup, 2), "years\n")
cat("STAR IQR:", round(star_iqr_followup[1], 2), "-", 
    round(star_iqr_followup[2], 2), "years\n\n")

# Summary table
followup_summary <- data.frame(
  Cohort = c("SRTR", "OPTN STAR"),
  `Median (years)` = c(round(srtr_median_followup, 2), 
                       round(star_median_followup, 2)),
  `IQR (years)` = c(paste0(round(srtr_iqr_followup[1], 2), "-", 
                           round(srtr_iqr_followup[2], 2)),
                    paste0(round(star_iqr_followup[1], 2), "-", 
                           round(star_iqr_followup[2], 2)))
)

print(followup_summary)
```

# STAR Status Justification Analysis (RECONSTITUTION OF RESULTS)

```{r star_justification}
# ============================================================================
# BUILD STAR STATUS JUSTIFICATION COHORT
# ============================================================================

star_lvad_status4 <- thoracic_stat4_2025q2 %>%
  dplyr::select(WL_ID_CODE, PRIM_DEVICETYPEID, PRIM_IMPLANTDT, PRIM_VADBRANDID) %>%
  filter(PRIM_DEVICETYPEID == 12)

star_lvad_status3 <- thoracic_stat3_2025q2 %>%
  dplyr::select(WL_ID_CODE, PRIM_DEVICETYPEID, PRIM_IMPLANTDT, PRIM_VADBRANDID) %>%
  filter(PRIM_DEVICETYPEID == 12)

star_lvad_status2 <- thoracic_stat2_2025q2 %>%
  dplyr::select(WL_ID_CODE, PRIM_DEVICETYPEID, PRIM_IMPLANTDT, PRIM_VADBRANDID) %>%
  filter(PRIM_DEVICETYPEID == 12)

star_lvad_status1 <- thoracic_stat1_2025q2 %>%
  dplyr::select(WL_ID_CODE, PRIM_DEVICETYPEID, PRIM_IMPLANTDT, PRIM_VADBRANDID) %>%
  filter(PRIM_DEVICETYPEID == 12)

star_lvads_from_justifications <- rbind(star_lvad_status4, star_lvad_status3,
                                        star_lvad_status2, star_lvad_status1) %>%
  distinct(WL_ID_CODE, PRIM_IMPLANTDT, .keep_all = TRUE)

star_lvads_just <- star_lvads_from_justifications %>%
  group_by(WL_ID_CODE) %>%
  arrange(PRIM_IMPLANTDT) %>%
  summarize(
    first_lvad_implant = first(as.Date(PRIM_IMPLANTDT, origin = "1970-01-01")),
    .groups = "drop"
  )

cat("STAR (Justification): Total patients with LVAD:", nrow(star_lvads_just), "\n")

# ============================================================================
# PREVALENCE ANALYSIS
# ============================================================================

get_star_just_lvad_prevalence <- function(check_date) {
  candidates_on_date <- thoracic_data_2025q2 %>%
    filter(INIT_DATE <= check_date) %>%
    filter(is.na(END_DATE) | END_DATE > check_date) %>%
    filter(INIT_AGE > 17) %>%
    dplyr::select(WL_ID_CODE)
  
  candidates_with_lvad <- candidates_on_date %>%
    left_join(star_lvads_just %>% dplyr::select(WL_ID_CODE, first_lvad_implant),
              by = "WL_ID_CODE") %>%
    mutate(has_lvad = !is.na(first_lvad_implant) & first_lvad_implant <= check_date)
  
  return(data.frame(
    date = check_date,
    total_waitlist = nrow(candidates_on_date),
    total_with_lvad = sum(candidates_with_lvad$has_lvad, na.rm = TRUE)
  ))
}

star_just_prevalence_list <- lapply(dates_to_check, get_star_just_lvad_prevalence)
star_just_prevalence_df <- bind_rows(star_just_prevalence_list) %>%
  mutate(
    lvad_proportion = total_with_lvad / total_waitlist * 100,
    date_label = format(date, "%b %d, %Y"),
    date_factor = factor(date_label, levels = date_label)
  )
```

# Three-Way Comparison: SRTR vs STAR (Justification) vs STAR (MCS)

```{r three_way_comparison}
# ============================================================================
# COMBINE ALL THREE SOURCES
# ============================================================================

combined_prevalence_3way <- bind_rows(
  prevalence_results_df %>%
    dplyr::select(date, total_waitlist, total_with_lvad, lvad_proportion, date_label) %>%
    mutate(data_source = "SRTR (MCS File)"),
  star_just_prevalence_df %>%
    dplyr::select(date, total_waitlist, total_with_lvad, lvad_proportion, date_label) %>%
    mutate(data_source = "STAR (MCS + Status Just. Files)"),
  star_prevalence_df %>%
    dplyr::select(date, total_waitlist, total_with_lvad, lvad_proportion, date_label) %>%
    mutate(data_source = "STAR (MCS File)")
) %>%
  mutate(
    date_factor = factor(date_label, levels = unique(date_label)),
    data_source = factor(data_source,
                         levels = c("SRTR (MCS File)", "STAR (MCS + Status Just. Files)", "STAR (MCS File)"))
  )

# ============================================================================
# PLOTS
# ============================================================================

prev_comparison_lvad_3way <- ggplot(combined_prevalence_3way,
                                    aes(x = date_factor, group = data_source)) +
  geom_line(aes(y = total_with_lvad, color = data_source, linetype = data_source),
            linewidth = 1.2) +
  geom_point(aes(y = total_with_lvad, color = data_source, shape = data_source),
             size = 2.5) +
  scale_color_manual(values = c("SRTR (MCS File)" = "coral",
                                "STAR (MCS + Status Just. Files)" = "darkgreen",
                                "STAR (MCS File)" = "steelblue")) +
  scale_linetype_manual(values = c("SRTR (MCS File)" = "solid",
                                   "STAR (MCS + Status Just. Files)" = "solid",
                                   "STAR (MCS File)" = "dashed")) +
  scale_shape_manual(values = c("SRTR (MCS File)" = 16,
                                "STAR (MCS + Status Just. Files)" = 15,
                                "STAR (MCS File)" = 17)) +
  scale_y_continuous(labels = comma, limits = c(0, NA)) +
  labs(x = "", y = "Number of Candidates\nwith LVADs",
       color = "Data Source", linetype = "Data Source", shape = "Data Source") +
  theme_classic(base_size = 15) +
  theme(
    legend.position = "bottom",
    axis.text.x = element_text(angle = 45, hjust = 1, size = 15),
    axis.text.y = element_text(size = 15),
    axis.title = element_text(size = 15),
    legend.text = element_text(size = 15)
  )

prev_comparison_proportion_3way <- ggplot(combined_prevalence_3way,
                                          aes(x = date_factor, y = lvad_proportion,
                                              fill = data_source)) +
  geom_col(position = "dodge", alpha = 1) +
  scale_fill_manual(values = c("SRTR (MCS File)" = "coral",
                               "STAR (MCS + Status Just. Files)" = "darkgreen",
                               "STAR (MCS File)" = "steelblue")) +
  scale_y_continuous(limits = c(0, NA)) +
  labs(x = "Date", y = "% of Waitlist\nwith LVADs", fill = "Data Source") +
  theme_classic(base_size = 15) +
  theme(
    legend.position = "bottom",
    axis.text.x = element_text(angle = 45, hjust = 1, size = 15),
    axis.text.y = element_text(size = 15),
    axis.title = element_text(size = 15),
    legend.text = element_text(size = 15)
  )

combined_prev_3way_plot <- (prev_comparison_lvad_3way + ggtitle("A")) /
  (prev_comparison_proportion_3way + ggtitle("B")) +
  plot_layout(guides = "collect") &
  theme(
    legend.position = "bottom",
    plot.title = element_text(size = 18, face = "bold", hjust = 0)
  ) &
  guides(
    color = guide_legend(nrow = 2),
    linetype = guide_legend(nrow = 2),
    shape = guide_legend(nrow = 2),
    fill = guide_legend(nrow = 2)
  )

print(combined_prev_3way_plot)

# Summary stats
cat("\n=== THREE-WAY PREVALENCE COMPARISON ===\n")
cat("SRTR at Oct 18, 2018:", prevalence_results_df$total_with_lvad[1], "\n")
cat("STAR (Justification) at Oct 18, 2018:", star_just_prevalence_df$total_with_lvad[1], "\n")
cat("STAR (MCS File) at Oct 18, 2018:", star_prevalence_df$total_with_lvad[1], "\n")
```

# Three-Way Outcomes: SRTR vs STAR (Justification) vs STAR (MCS)

```{r outcomes_3way}
star_just_lvad_outcomes <- star_lvads_just %>%
  filter(first_lvad_implant >= mdy("10-18-2018") &
         first_lvad_implant <= mdy("12-31-2024")) %>%
  left_join(thoracic_data_2025q2 %>%
              dplyr::select(WL_ID_CODE, INIT_DATE, END_DATE, REM_CD, INIT_AGE),
            by = "WL_ID_CODE") %>%
  filter(INIT_AGE > 17) %>%
  mutate(
    removal_or_censor_date = if_else(is.na(END_DATE),
                                     as.Date("2025-06-30"),
                                     as.Date(END_DATE)),
    end_date = removal_or_censor_date,
    followup_time_original = as.numeric(difftime(end_date, first_lvad_implant, units = "days")) / 365.25,
    followup_time = pmin(followup_time_original, 3),
    event = case_when(
      followup_time_original > 3 ~ 0,  # Censor if event occurred after 3 years
      REM_CD %in% c(8, 13) ~ 1,
      REM_CD == 4 ~ 2,
      TRUE ~ 0
    ),
    event_factor = factor(event, levels = c(0, 1, 2),
                          labels = c("Censored", "Death/Deterioration", "Transplant")),
    data_source = "STAR (MCS + Status Just. Files)"
  ) %>%
  filter(followup_time > 0)

cat("\nSTAR (Justification) outcomes cohort:", nrow(star_just_lvad_outcomes), "patients\n")
cat("  Deaths/Deterioration:", sum(star_just_lvad_outcomes$event == 1), "\n")
cat("  Transplants:", sum(star_just_lvad_outcomes$event == 2), "\n")
cat("  Censored:", sum(star_just_lvad_outcomes$event == 0), "\n")

# ============================================================================
# COMBINE SRTR AND STAR (JUSTIFICATION)
# ============================================================================

combined_cohort_2way <- bind_rows(
  srtr_lvad_outcomes %>% dplyr::select(followup_time, event_factor, data_source),
  star_just_lvad_outcomes %>% dplyr::select(followup_time, event_factor, data_source)
) %>%
  mutate(data_source = factor(data_source, levels = c("SRTR (MCS File)", "STAR (MCS + Status Just. Files)")))

# Cumulative incidence
cif_fit_2way <- tidycmprsk::cuminc(Surv(followup_time, event_factor) ~ data_source,
                                    data = combined_cohort_2way)

# Fine-Gray tests
fg_transplant <- tidycmprsk::crr(Surv(followup_time, event_factor) ~ data_source,
                                  data = combined_cohort_2way,
                                  failcode = "Transplant")
fg_death <- tidycmprsk::crr(Surv(followup_time, event_factor) ~ data_source,
                             data = combined_cohort_2way,
                             failcode = "Death/Deterioration")

p_transplant <- tidy(fg_transplant)$p.value[1]
p_death <- tidy(fg_death)$p.value[1]

cat("\n=== FINE-GRAY TESTS: SRTR vs STAR (JUSTIFICATION) ===\n")
cat("Transplant p:", format.pval(p_transplant, digits = 3), "\n")
cat("Death/Deterioration p:", format.pval(p_death, digits = 3), "\n")

# ============================================================================
# PLOT
# ============================================================================

cif_plot_2way <- cif_fit_2way %>%
  ggcuminc(
    outcome = c("Death/Deterioration", "Transplant"),
    linewidth = 1.2
  ) +
  add_confidence_interval(alpha = 0.2) +
  add_risktable(
    size = 5,
    theme = theme_risktable_default(
      plot.title.size = 14,
      axis.text.y.size = 14
    )
  ) +
  scale_x_continuous(breaks = 0:3, limits = c(0, 3)) +
  scale_y_continuous(
    limits = c(0, 1),
    labels = scales::percent,
    expand = c(0.01, 0)
  ) +
  xlab("Years from LVAD Implantation") +
  ylab("Cumulative Incidence") +
  scale_color_manual(
    values = c("SRTR (MCS File)" = "coral", "STAR (MCS + Status Just. Files)" = "darkgreen"),
    name = "Data Source"
  ) +
  scale_fill_manual(
    values = c("SRTR (MCS File)" = "coral", "STAR (MCS + Status Just. Files)" = "darkgreen"),
    name = "Data Source"
  ) +
  scale_linetype_discrete(name = "Outcome") +
  guides(
    color = guide_legend(order = 1, nrow = 1),
    fill = guide_legend(order = 1, nrow = 1),
    linetype = guide_legend(order = 2, nrow = 1)
  ) +
  # annotate("text", x = 2.8, y = 0.95,
  #          label = paste0("Fine-Gray Transplant: ", fmt_p(p_transplant)),
  #          hjust = 1, size = 5, fontface = "bold") +
  # annotate("text", x = 2.8, y = 0.90,
  #          label = paste0("Fine-Gray Death/Deterioration: ", fmt_p(p_death)),
  #          hjust = 1, size = 5, fontface = "bold") +
  theme_classic() +
  theme(
    legend.position = "bottom",
    legend.box = "vertical",
    legend.title = element_text(size = 15, face = "bold"),
    axis.text = element_text(size = 15),
    axis.title = element_text(size = 15),
    legend.text = element_text(size = 15)
  )

print(cif_plot_2way)

# Summary stats
cif_summary_2way <- cif_fit_2way %>%
  tidy() %>%
  filter(time <= 3.5) %>%
  group_by(outcome, strata) %>%
  filter(abs(time - 1) == min(abs(time - 1)) |
           abs(time - 2) == min(abs(time - 2)) |
           abs(time - 3) == min(abs(time - 3))) %>%
  distinct(outcome, strata, .keep_all = TRUE) %>%
  arrange(outcome, strata, time) %>%
  mutate(
    time_year = round(time, 1),
    estimate_pct = round(estimate * 100, 1)
  ) %>%
  dplyr::select(outcome, strata, time_year, estimate_pct)

cat("\n=== CUMULATIVE INCIDENCE AT 1, 2, 3 YEARS ===\n")
print(cif_summary_2way)
```

# IABP Generalizability Analysis

```{r iabp_analysis}
# ============================================================================
# BUILD IABP COHORTS
# ============================================================================

# SRTR IABP
srtr_iabp <- thor_support_device %>%
  filter(DeviceTypeId == 8) %>%
  group_by(PX_ID) %>%
  arrange(ImplantDt) %>%
  slice(1) %>%
  ungroup() %>%
  dplyr::select(PX_ID, iabp_implant_date = ImplantDt)

# STAR IABP (Status Justification)
star_iabp_status4 <- thoracic_stat4_2025q2 %>%
  dplyr::select(WL_ID_CODE, PRIM_DEVICETYPEID, PRIM_IMPLANTDT) %>%
  filter(PRIM_DEVICETYPEID == 8)

star_iabp_status3 <- thoracic_stat3_2025q2 %>%
  dplyr::select(WL_ID_CODE, PRIM_DEVICETYPEID, PRIM_IMPLANTDT) %>%
  filter(PRIM_DEVICETYPEID == 8)

star_iabp_status2 <- thoracic_stat2_2025q2 %>%
  dplyr::select(WL_ID_CODE, PRIM_DEVICETYPEID, PRIM_IMPLANTDT) %>%
  filter(PRIM_DEVICETYPEID == 8)

star_iabp_status1 <- thoracic_stat1_2025q2 %>%
  dplyr::select(WL_ID_CODE, PRIM_DEVICETYPEID, PRIM_IMPLANTDT) %>%
  filter(PRIM_DEVICETYPEID == 8)

star_iabp_from_justifications <- rbind(star_iabp_status4, star_iabp_status3,
                                       star_iabp_status2, star_iabp_status1) %>%
  distinct(WL_ID_CODE, PRIM_IMPLANTDT, .keep_all = TRUE) %>%
  filter(PRIM_IMPLANTDT <= as.Date("2024-12-31"))

star_iabp_just <- star_iabp_from_justifications %>%
  group_by(WL_ID_CODE) %>%
  arrange(PRIM_IMPLANTDT) %>%
  summarize(
    first_iabp_implant = first(as.Date(PRIM_IMPLANTDT, origin = "1970-01-01")),
    .groups = "drop"
  )

# STAR IABP (MCS File)
star_iabp <- thoracic_mcs_device_2025q2 %>%
  filter(DEVICE_TYPE == "IABP") %>%
  group_by(WL_ID_CODE) %>%
  arrange(IMPLANT_DATE) %>%
  summarize(
    first_iabp_implant = first(IMPLANT_DATE),
    .groups = "drop"
  )

# ============================================================================
# PREVALENCE ANALYSIS (JAN 2020 - APRIL 2025)
# ============================================================================

dates_to_check_iabp <- seq(mdy("01-01-2020"), mdy("01-01-2025"), by = "2 months")

get_srtr_iabp_prevalence <- function(check_date) {
  candidates_on_date <- cand_thor2025q2 %>%
    filter(CAN_LISTING_DT <= check_date) %>%
    filter(is.na(CAN_REM_DT) | CAN_REM_DT > check_date) %>%
    filter(CAN_AGE_IN_MONTHS_AT_LISTING / 12 > 17) %>%
    dplyr::select(PX_ID)
  
  candidates_with_iabp <- candidates_on_date %>%
    left_join(srtr_iabp %>% dplyr::select(PX_ID, iabp_implant_date), by = "PX_ID") %>%
    mutate(has_iabp = !is.na(iabp_implant_date) & iabp_implant_date <= check_date)
  
  return(data.frame(
    date = check_date,
    total_waitlist = nrow(candidates_on_date),
    total_with_device = sum(candidates_with_iabp$has_iabp, na.rm = TRUE)
  ))
}

get_star_iabp_just_prevalence <- function(check_date) {
  candidates_on_date <- thoracic_data_2025q2 %>%
    filter(INIT_DATE <= check_date) %>%
    filter(is.na(END_DATE) | END_DATE > check_date) %>%
    filter(INIT_AGE > 17) %>%
    dplyr::select(WL_ID_CODE)
  
  candidates_with_iabp <- candidates_on_date %>%
    left_join(star_iabp_just %>% dplyr::select(WL_ID_CODE, first_iabp_implant),
              by = "WL_ID_CODE") %>%
    mutate(has_iabp = !is.na(first_iabp_implant) & first_iabp_implant <= check_date)
  
  return(data.frame(
    date = check_date,
    total_waitlist = nrow(candidates_on_date),
    total_with_device = sum(candidates_with_iabp$has_iabp, na.rm = TRUE)
  ))
}

get_star_iabp_prevalence <- function(check_date) {
  candidates_on_date <- thoracic_data_2025q2 %>%
    filter(INIT_DATE <= check_date) %>%
    filter(is.na(END_DATE) | END_DATE > check_date) %>%
    filter(INIT_AGE > 17) %>%
    dplyr::select(WL_ID_CODE)
  
  candidates_with_iabp <- candidates_on_date %>%
    left_join(star_iabp %>% dplyr::select(WL_ID_CODE, first_iabp_implant), 
              by = "WL_ID_CODE") %>%
    mutate(has_iabp = !is.na(first_iabp_implant) & first_iabp_implant <= check_date)
  
  return(data.frame(
    date = check_date,
    total_waitlist = nrow(candidates_on_date),
    total_with_device = sum(candidates_with_iabp$has_iabp, na.rm = TRUE)
  ))
}

# Run analyses
srtr_iabp_prevalence_list <- lapply(dates_to_check_iabp, get_srtr_iabp_prevalence)
srtr_iabp_prevalence_df <- bind_rows(srtr_iabp_prevalence_list) %>%
  mutate(
    device_proportion = total_with_device / total_waitlist * 100,
    date_label = format(date, "%b %d, %Y"),
    date_factor = factor(date_label, levels = date_label)
  )

star_iabp_just_prevalence_list <- lapply(dates_to_check_iabp, get_star_iabp_just_prevalence)
star_iabp_just_prevalence_df <- bind_rows(star_iabp_just_prevalence_list) %>%
  mutate(
    device_proportion = total_with_device / total_waitlist * 100,
    date_label = format(date, "%b %d, %Y"),
    date_factor = factor(date_label, levels = date_label)
  )

star_iabp_prevalence_list <- lapply(dates_to_check_iabp, get_star_iabp_prevalence)
star_iabp_prevalence_df <- bind_rows(star_iabp_prevalence_list) %>%
  mutate(
    device_proportion = total_with_device / total_waitlist * 100,
    date_label = format(date, "%b %d, %Y"),
    date_factor = factor(date_label, levels = date_label)
  )

# ============================================================================
# PLOT
# ============================================================================

iabp_comparison <- bind_rows(
  srtr_iabp_prevalence_df %>%
    dplyr::select(date, total_with_device, date_label) %>%
    mutate(data_source = "SRTR (MCS File)"),
  star_iabp_just_prevalence_df %>%
    dplyr::select(date, total_with_device, date_label) %>%
    mutate(data_source = "STAR (MCS + Status Just. Files)"),
  star_iabp_prevalence_df %>%
    dplyr::select(date, total_with_device, date_label) %>%
    mutate(data_source = "STAR (MCS File)")
) %>%
  mutate(
    date_factor = factor(date_label, levels = unique(date_label)),
    data_source = factor(data_source, 
                         levels = c("SRTR (MCS File)", "STAR (MCS + Status Just. Files)", "STAR (MCS File)"))
  )

iabp_3way_plot <- ggplot(iabp_comparison,
                         aes(x = date_factor, y = total_with_device,
                             color = data_source, linetype = data_source,
                             group = data_source)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2.5) +
  scale_color_manual(values = c("SRTR (MCS File)" = "coral",
                                "STAR (MCS + Status Just. Files)" = "darkgreen",
                                "STAR (MCS File)" = "steelblue")) +
  scale_linetype_manual(values = c("SRTR (MCS File)" = "solid",
                                   "STAR (MCS + Status Just. Files)" = "solid",
                                   "STAR (MCS File)" = "dashed")) +
  scale_y_continuous(labels = comma, limits = c(0, NA)) +
  labs(
    x = "Date",
    y = "Number of Candidates with IABP",
    color = "Data Source",
    linetype = "Data Source"
  ) +
  theme_classic(base_size = 15) +
  theme(
    legend.position = "bottom",
    axis.text.x = element_text(angle = 45, hjust = 1, size = 15),
    axis.text.y = element_text(size = 15),
    axis.title = element_text(size = 15),
    legend.text = element_text(size = 15)
  )

print(iabp_3way_plot)




iabp_comparison_table <- iabp_comparison %>%
  select(date, date_label, data_source, total_with_device) %>%
  tidyr::pivot_wider(
    names_from = data_source,
    values_from = total_with_device
  ) %>%
  mutate(
    Date = date_label,
    `Absolute Difference (SRTR vs STAR MCS)` = `SRTR (MCS File)` - `STAR (MCS File)`
  ) %>%
  select(Date, `SRTR (MCS File)`, `STAR (Status Justification)`, `STAR (MCS File)`, 
         `Absolute Difference (SRTR vs STAR MCS)`)

# Create formatted table
iabp_table <- kable(iabp_comparison_table, 
                    format = "html",
                    caption = "Supplemental Table: IABP Prevalence Comparison",
                    align = c('l', 'r', 'r', 'r', 'r')) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE, 
                font_size = 11)

print(iabp_table)





```

# Status 4 LVAD Analysis
```{r status4_analysis}
library(dplyr)
library(ggplot2)
library(lubridate)
library(tidyr)

# Define dates to check - LIMITED TO APRIL 30, 2025
dates_to_check <- c(
  mdy("10-18-2018"),
  seq(mdy("01-01-2019"), mdy("05-01-2025"), by = "2 months") - days(1)
)

# ============================================================================
# SRTR STATUS 4 ANALYSIS
# ============================================================================

# Prepare LVAD status 4 justification data
lvadstatus4 <- just_form_hr_stat4 %>% 
  select(PX_ID, CriteriaLvadSupport, FormEffectiveDt, FormExpirationDt, RequestedCandStatCd) %>% 
  filter(CriteriaLvadSupport == 1)

# Function to get status 4 LVAD counts at each date
get_srtr_status4_lvad <- function(check_date) {
  
  cat("Processing SRTR date:", format(check_date, "%Y-%m-%d"), "\n")
  
  # Get all candidates actively listed on this date
  candidates_on_date <- cand_thor2025q2 %>%
    filter(CAN_LISTING_DT <= check_date) %>%
    filter(is.na(CAN_REM_DT) | CAN_REM_DT > check_date) %>%
    filter(CAN_AGE_IN_MONTHS_AT_LISTING / 12 > 17) %>%
    select(PX_ID, CAN_LISTING_DT, CAN_REM_DT)
  
  # Merge with status history to get current status
  candidates_with_status <- candidates_on_date %>%
    left_join(stathist_thor2025q2, by = "PX_ID") %>%
    filter(CANHX_BEGIN_DT <= check_date) %>%
    filter(is.na(CANHX_END_DT) | CANHX_END_DT > check_date) %>%
    group_by(PX_ID) %>%
    arrange(desc(CANHX_BEGIN_DT)) %>%
    slice(1) %>%
    ungroup()
  
  # Filter to status 4 candidates (status code 2140)
  status4_candidates <- candidates_with_status %>%
    filter(CANHX_STAT_CD == 2140)
  
  # Semi-join by ID only - no date filtering to avoid join artifacts
  status4_with_lvad_justification <- status4_candidates %>%
    semi_join(lvadstatus4, by = "PX_ID")
  
  total_status4 <- nrow(status4_candidates)
  status4_with_lvad <- nrow(status4_with_lvad_justification)
  
  return(data.frame(
    date = check_date,
    total_status4 = total_status4,
    status4_with_lvad = status4_with_lvad,
    proportion = ifelse(total_status4 > 0, status4_with_lvad / total_status4 * 100, 0)
  ))
}

# Run analysis
status4_results_list <- lapply(dates_to_check, get_srtr_status4_lvad)
status4_results_df <- bind_rows(status4_results_list)

status4_results_df <- status4_results_df %>%
  mutate(
    date_label = format(date, "%b %d, %Y"),
    date_factor = factor(date_label, levels = date_label)
  )

cat("\n=== STATUS 4 CANDIDATES WITH LVAD SUPPORT (SRTR) ===\n\n")
print(status4_results_df)

# ============================================================================
# COMBINE WITH SRTR PREVALENCE DATA
# ============================================================================

combined_plot_data <- status4_results_df %>%
  left_join(
    prevalence_results_df %>% select(date, total_with_lvad),
    by = "date"
  )

# SRTR combined plot
srtr_status4_plot <- ggplot(combined_plot_data, aes(x = date_factor)) +
  geom_col(aes(y = total_status4, fill = "All Candidates with Status 4 Justifications"), alpha = 0.5) +
  geom_col(aes(y = status4_with_lvad, fill = "Candidates with Status 4 Justifications for LVAD")) +
  geom_line(aes(y = total_with_lvad, color = "All Candidates with LVAD per Device File", group = 1), 
            linewidth = 1.5) +
  geom_point(aes(y = total_with_lvad, color = "All Candidates with LVAD per Device File"), 
             size = 2.5) +
  scale_fill_manual(
    name = "",
    values = c("All Candidates with Status 4 Justifications" = "#0072B2",
               "Candidates with Status 4 Justifications for LVAD" = "#D55E00"),
    breaks = c("All Candidates with Status 4 Justifications", 
               "Candidates with Status 4 Justifications for LVAD")
  ) +
  scale_color_manual(
    name = "",
    values = c("All Candidates with LVAD per Device File" = "#CC0000")
  ) +
  labs(
    title = "Status 4 Candidates and Overall LVAD Prevalence (SRTR)",
    x = "Date",
    y = "Number of Candidates"
  ) +
  theme_classic(base_size = 15) +
  theme(
    plot.title = element_text(face = "bold", size = 15),
    axis.title = element_text(face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 15),
    legend.position = "top",
    legend.text = element_text(size = 15)
  ) +
  scale_y_continuous(labels = scales::comma) +
  guides(
    fill = guide_legend(order = 1),
    color = guide_legend(order = 2)
  )

print(srtr_status4_plot)

# ============================================================================
# STAR STATUS 4 ANALYSIS
# ============================================================================

lvadstatus4star <- thoracic_stat4_2025q2 %>% 
  select(WL_ID_CODE, CRITERIALVADSUPPORT, FORMEFFECTIVEDT, UNOS_CAND_STAT_CD) %>% 
  filter(CRITERIALVADSUPPORT == 1) %>%
  mutate(new_UNOS_CAND_STAT_CD = ifelse(UNOS_CAND_STAT_CD == 2020, 2140, UNOS_CAND_STAT_CD))

star_waitlist_hx <- thoracic_wlhistory_data_2025q2 %>% 
  select(WL_ID_CODE, CHG_TY, CHG_DATE, UNOS_CAND_STAT_CD)

# Function to get status 4 LVAD counts at each date for STAR
get_star_status4_lvad <- function(check_date) {
  
  cat("Processing STAR date:", format(check_date, "%Y-%m-%d"), "\n")
  
  candidates_on_date <- thoracic_data_2025q2 %>%
    filter(INIT_DATE <= check_date) %>%
    filter(is.na(END_DATE) | END_DATE > check_date) %>%
    filter(INIT_AGE > 17) %>%
    select(WL_ID_CODE, INIT_DATE, END_DATE)
  
  status_history_with_end <- star_waitlist_hx %>%
    group_by(WL_ID_CODE) %>%
    arrange(CHG_DATE) %>%
    mutate(status_end_date = lead(CHG_DATE)) %>%
    ungroup()
  
  candidates_with_status <- candidates_on_date %>%
    left_join(status_history_with_end, by = "WL_ID_CODE") %>%
    filter(CHG_DATE <= check_date) %>%
    filter(is.na(status_end_date) | status_end_date > check_date) %>%
    group_by(WL_ID_CODE) %>%
    arrange(desc(CHG_DATE)) %>%
    slice(1) %>%
    ungroup()
  
  status4_candidates <- candidates_with_status %>%
    filter(UNOS_CAND_STAT_CD == 2140)
  
  # Semi-join by ID only - no date filtering to avoid join artifacts
  status4_with_lvad_justification <- status4_candidates %>%
    semi_join(lvadstatus4star, by = "WL_ID_CODE")
  
  total_status4 <- nrow(status4_candidates)
  status4_with_lvad <- nrow(status4_with_lvad_justification)
  
  return(data.frame(
    date = check_date,
    total_status4 = total_status4,
    status4_with_lvad = status4_with_lvad,
    proportion = ifelse(total_status4 > 0, status4_with_lvad / total_status4 * 100, 0)
  ))
}

# Run analysis
star_status4_results_list <- lapply(dates_to_check, get_star_status4_lvad)
star_status4_results_df <- bind_rows(star_status4_results_list)

star_status4_results_df <- star_status4_results_df %>%
  mutate(
    date_label = format(date, "%b %d, %Y"),
    date_factor = factor(date_label, levels = date_label)
  )

cat("\n=== STATUS 4 CANDIDATES WITH LVAD SUPPORT (STAR) ===\n\n")
print(star_status4_results_df)

# ============================================================================
# COMBINE WITH STAR PREVALENCE DATA
# ============================================================================

star_combined_plot_data <- star_status4_results_df %>%
  left_join(
    star_prevalence_df %>% select(date, total_with_lvad),
    by = "date"
  )

# STAR combined plot
star_status4_plot <- ggplot(star_combined_plot_data, aes(x = date_factor)) +
  geom_col(aes(y = total_status4, fill = "All Candidates with Status 4 Justifications"), alpha = 0.5) +
  geom_col(aes(y = status4_with_lvad, fill = "Candidates with Status 4 Justifications for LVAD")) +
  geom_line(aes(y = total_with_lvad, color = "All Candidates with LVAD per MCS File", group = 1), 
            linewidth = 1.5) +
  geom_point(aes(y = total_with_lvad, color = "All Candidates with LVAD per MCS File"), 
             size = 2.5) +
  scale_fill_manual(
    name = "",
    values = c("All Candidates with Status 4 Justifications" = "#0072B2",
               "Candidates with Status 4 Justifications for LVAD" = "#D55E00"),
    breaks = c("All Candidates with Status 4 Justifications", 
               "Candidates with Status 4 Justifications for LVAD")
  ) +
  scale_color_manual(
    name = "",
    values = c("All Candidates with LVAD per MCS File" = "#CC0000")
  ) +
  labs(
    title = "Status 4 Candidates and Overall LVAD Prevalence (OPTN STAR)",
    x = "Date",
    y = "Number of Candidates"
  ) +
  theme_classic(base_size = 15) +
  theme(
    plot.title = element_text(face = "bold", size = 15),
    axis.title = element_text(face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 15),
    legend.position = "top",
    legend.text = element_text(size = 15)
  ) +
  scale_y_continuous(labels = scales::comma) +
  guides(
    fill = guide_legend(order = 1),
    color = guide_legend(order = 2)
  )

print(star_status4_plot)

# ============================================================================
# COMPARISON PLOT (SRTR vs STAR)
# ============================================================================

srtr_for_comparison <- combined_plot_data %>%
  mutate(data_source = "SRTR")

star_for_comparison <- star_combined_plot_data %>%
  mutate(data_source = "STAR")

both_sources <- bind_rows(srtr_for_comparison, star_for_comparison)

comparison_plot <- ggplot(both_sources, aes(x = date_factor)) +
  geom_col(aes(y = total_status4, fill = "All Candidates with Status 4 Justifications"), 
           alpha = 0.7) +
  geom_col(aes(y = status4_with_lvad, fill = "Candidates with Status 4 Justifications for LVAD")) +
  geom_line(aes(y = total_with_lvad, color = "All Candidates with LVAD per MCS File", group = 1), 
            linewidth = 1.2) +
  geom_point(aes(y = total_with_lvad, color = "All Candidates with LVAD per MCS File"), 
             size = 2) +
  scale_fill_manual(
    name = "",
    values = c("All Candidates with Status 4 Justifications" = "steelblue",
               "Candidates with Status 4 Justifications for LVAD" = "coral"),
    breaks = c("All Candidates with Status 4 Justifications", 
               "Candidates with Status 4 Justifications for LVAD")
  ) +
  scale_color_manual(
    name = "",
    values = c("All Candidates with LVAD per MCS File" = "#b24745")
  ) +
  facet_wrap(~data_source, ncol = 1, scales = "free_y") +
  labs(
    x = "Date",
    y = "Number of Candidates"
  ) +
  theme_classic(base_size = 15) +
  theme(
    axis.title = element_text(face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 15),
    axis.text.y = element_text(size = 15),
    legend.position = "top",
    legend.text = element_text(size = 15),
    strip.text = element_text(face = "bold", size = 15)
  ) +
  scale_y_continuous(labels = scales::comma) +
  guides(
    fill = guide_legend(order = 1),
    color = guide_legend(order = 2)
  )

print(comparison_plot)
```


```{r supplemental_tables}
library(knitr)
library(kableExtra)

# ============================================================================
# SRTR STATUS 4 AND PREVALENCE TABLE
# ============================================================================

srtr_status_prevalence_table <- combined_plot_data %>%
  select(
    date,
    date_label,
    total_status4,
    status4_with_lvad,
    total_with_lvad
  ) %>%
  mutate(
    Date = date_label,
    `Total Status 4` = total_status4,
    `Status 4 with LVAD` = status4_with_lvad,
    `Total LVAD Prevalence` = total_with_lvad
  ) %>%
  select(Date, `Total Status 4`, `Status 4 with LVAD`, `Total LVAD Prevalence`)

cat("\n=== SRTR STATUS 4 AND PREVALENCE TABLE ===\n\n")
print(srtr_status_prevalence_table)

# Create formatted HTML table
srtr_status_html <- kable(srtr_status_prevalence_table, format = "html", 
                          caption = "Table: SRTR Status 4 Composition and LVAD Prevalence") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), 
                full_width = FALSE, font_size = 11) %>%
  add_header_above(c(" " = 1, "Status 4 Composition" = 2, "Prevalence" = 1))

# ============================================================================
# STAR STATUS 4 AND PREVALENCE TABLE
# ============================================================================

star_status_prevalence_table <- star_combined_plot_data %>%
  select(
    date,
    date_label,
    total_status4,
    status4_with_lvad,
    total_with_lvad
  ) %>%
  mutate(
    Date = date_label,
    `Total Status 4` = total_status4,
    `Status 4 with LVAD` = status4_with_lvad,
    `Total LVAD Prevalence` = total_with_lvad
  ) %>%
  select(Date, `Total Status 4`, `Status 4 with LVAD`, `Total LVAD Prevalence`)

cat("\n=== STAR STATUS 4 AND PREVALENCE TABLE ===\n\n")
print(star_status_prevalence_table)

# Create formatted HTML table
star_status_html <- kable(star_status_prevalence_table, format = "html",
                          caption = "Table: STAR Status 4 Composition and LVAD Prevalence") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE, font_size = 11) %>%
  add_header_above(c(" " = 1, "Status 4 Composition" = 2, "Prevalence" = 1))

# ============================================================================
# SRTR FLOW TABLE
# ============================================================================

srtr_flow_table <- flow_results_df %>%
  mutate(
    Period = period_label,
    `New Listings with LVAD` = new_listings_with_lvad,
    `LVAD Implants on Waitlist` = lvad_implants_on_waitlist
  ) %>%
  select(Period, `New Listings with LVAD`, `LVAD Implants on Waitlist`)

cat("\n=== SRTR FLOW TABLE ===\n\n")
print(srtr_flow_table)

# Create formatted HTML table
srtr_flow_html <- kable(srtr_flow_table, format = "html",
                        caption = "Table: SRTR Waitlist Flow Dynamics") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE, font_size = 11)

# ============================================================================
# STAR FLOW TABLE
# ============================================================================

star_flow_table <- star_flow_df %>%
  mutate(
    Period = period_label,
    `New Listings with LVAD` = new_listings_with_lvad,
    `LVAD Implants on Waitlist` = lvad_implants_on_waitlist
  ) %>%
  select(Period, `New Listings with LVAD`, `LVAD Implants on Waitlist`)

cat("\n=== STAR FLOW TABLE ===\n\n")
print(star_flow_table)

# Create formatted HTML table
star_flow_html <- kable(star_flow_table, format = "html",
                        caption = "Table: STAR Waitlist Flow Dynamics") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE, font_size = 11)

# ============================================================================
# SRTR PREVALENCE TABLE
# ============================================================================

srtr_prevalence_html <- prevalence_results_df %>%
  select(date_label, total_waitlist, total_with_lvad, lvad_proportion) %>%
  mutate(lvad_proportion = round(lvad_proportion, 2)) %>%
  kable(format = "html",
        col.names = c("Date", "Total Waitlist", "Total with LVAD", "LVAD %"),
        align = c('l', 'r', 'r', 'r'),
        caption = "Table: SRTR LVAD Prevalence") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE, font_size = 11)

# ============================================================================
# STAR PREVALENCE TABLE
# ============================================================================

star_prevalence_html <- star_prevalence_df %>%
  select(date_label, total_waitlist, total_with_lvad, lvad_proportion) %>%
  mutate(lvad_proportion = round(lvad_proportion, 2)) %>%
  kable(format = "html",
        col.names = c("Date", "Total Waitlist", "Total with LVAD", "LVAD %"),
        align = c('l', 'r', 'r', 'r'),
        caption = "Table: STAR LVAD Prevalence") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE, font_size = 11)

# ============================================================================
# SRTR COMBINED TABLE (Waitlist + Status 4 + LVAD Prevalence)
# ============================================================================

srtr_combined_table <- combined_plot_data %>%
  select(date, date_label, total_status4, status4_with_lvad, total_with_lvad) %>%
  left_join(
    prevalence_results_df %>%
      select(date, total_waitlist, lvad_proportion) %>%
      mutate(lvad_proportion = round(lvad_proportion, 2)),
    by = "date"
  ) %>%
  select(
    Date = date_label,
    `Total Waitlist` = total_waitlist,
    `Total Status 4` = total_status4,
    `Status 4 with LVAD` = status4_with_lvad,
    `Total LVAD Prevalence` = total_with_lvad,
    `LVAD %` = lvad_proportion
  )

srtr_combined_html <- kable(srtr_combined_table, format = "html",
                            caption = "Table: SRTR Status 4 Composition and LVAD Prevalence") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE, font_size = 11) %>%
  add_header_above(c(" " = 1, " " = 1, "Status 4 Composition" = 2, "LVAD Prevalence" = 2))

# ============================================================================
# STAR COMBINED TABLE (Waitlist + Status 4 + LVAD Prevalence)
# ============================================================================

star_combined_table <- star_combined_plot_data %>%
  select(date, date_label, total_status4, status4_with_lvad, total_with_lvad) %>%
  left_join(
    star_prevalence_df %>%
      select(date, total_waitlist, lvad_proportion) %>%
      mutate(lvad_proportion = round(lvad_proportion, 2)),
    by = "date"
  ) %>%
  select(
    Date = date_label,
    `Total Waitlist` = total_waitlist,
    `Total Status 4` = total_status4,
    `Status 4 with LVAD` = status4_with_lvad,
    `Total LVAD Prevalence` = total_with_lvad,
    `LVAD %` = lvad_proportion
  )

star_combined_html <- kable(star_combined_table, format = "html",
                            caption = "Table: STAR Status 4 Composition and LVAD Prevalence") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE, font_size = 11) %>%
  add_header_above(c(" " = 1, " " = 1, "Status 4 Composition" = 2, "LVAD Prevalence" = 2))
```


# Table 1: Baseline Characteristics (No Statistical Tests)
```{r table1}
library(gtsummary)
library(labelled)

# Define LVAD keywords
lvad_keywords <- c("heartmate", "hm", "lionheart", "arrow", "jarvik", "micromed", 
                   "debakey", "novacor", "evaheart", "worldheart", "levacor", 
                   "reliantheart", "hvad", "heartsaver")

# ============================================================================
# PREPARE SRTR TABLE 1 DATA
# ============================================================================

srtr_table1 <- srtr_lvad_outcomes %>%
  select(PX_ID, lvad_implant_date) %>%
  left_join(
    cand_thor2025q2 %>%
      select(PX_ID, CAN_GENDER, CAN_RACE, CAN_ETHNICITY_SRTR, 
             CAN_AGE_IN_MONTHS_AT_LISTING, CAN_BMI, CAN_ABO, 
             CAN_DGN, CAN_PRIMARY_PAY, CAN_LISTING_DT),
    by = "PX_ID"
  ) %>%
  left_join(
    thor_support_device %>%
      filter(DeviceTypeId == 12 & VentricularSupportId == 1) %>%
      group_by(PX_ID) %>%
      arrange(ImplantDt) %>%
      slice(1) %>%
      ungroup() %>%
      select(PX_ID, VadBrandId),
    by = "PX_ID"
  ) %>%
  mutate(
    data_source = "SRTR",
    age_at_listing = floor(CAN_AGE_IN_MONTHS_AT_LISTING / 12),
    age_at_lvad = floor(CAN_AGE_IN_MONTHS_AT_LISTING / 12) + 
      as.numeric(difftime(lvad_implant_date, CAN_LISTING_DT, units = "days")) / 365.25,
    age_at_lvad = floor(age_at_lvad),
    sex = ifelse(CAN_GENDER == "M", "Male", "Female"),
    sex = factor(sex, levels = c("Male", "Female")),
    race = case_when(
      is.na(CAN_RACE) & CAN_ETHNICITY_SRTR == "LATINO" ~ "Hispanic/Latino",
      CAN_RACE == 8 & CAN_ETHNICITY_SRTR == "NLATIN" ~ "White",
      CAN_RACE == 16 & CAN_ETHNICITY_SRTR == "NLATIN" ~ "Black",
      CAN_RACE == 64 & CAN_ETHNICITY_SRTR == "NLATIN" ~ "Asian",
      TRUE ~ "Other"),
    race = factor(race, levels = c("White", "Black", "Hispanic/Latino", "Asian", "Other")),
    bmi_categories = case_when(
      CAN_BMI < 18 ~ "Underweight",
      CAN_BMI >= 18 & CAN_BMI < 25 ~ "Normal",
      CAN_BMI >= 25 & CAN_BMI < 30 ~ "Overweight",
      CAN_BMI >= 30 ~ "Obese",
      TRUE ~ "Unknown"),
    bmi_categories = factor(bmi_categories, levels = c("Normal", "Underweight", "Overweight", "Obese", "Unknown")),
    blood_type = factor(
      case_when(
        CAN_ABO %in% c("A", "A1", "A2") ~ "A",
        CAN_ABO %in% c("A1B", "A2B") ~ "AB",
        TRUE ~ CAN_ABO)),
    diagnosis = case_when(
      CAN_DGN > 999 & CAN_DGN < 1007 ~ "Dilated Cardiomyopathy, Non-Ischemic",
      CAN_DGN == 1007 | CAN_DGN == 1200 ~ "Ischemic Cardiomyopathy",
      CAN_DGN > 1048 & CAN_DGN < 1100 ~ "Restrictive Cardiomyopathy",
      CAN_DGN == 1203 | (CAN_DGN >= 1205 & CAN_DGN <= 1207) ~ "Congenital Heart Disease",
      TRUE ~ "Other"),
    diagnosis = factor(diagnosis, levels = c("Dilated Cardiomyopathy, Non-Ischemic", "Ischemic Cardiomyopathy", 
                                             "Restrictive Cardiomyopathy", "Congenital Heart Disease", "Other")),
    payor = case_when(
      CAN_PRIMARY_PAY %in% c(2, 3, 4, 5, 6, 7, 13) ~ "Public",
      CAN_PRIMARY_PAY == 1 ~ "Private",
      TRUE ~ "Other"),
    payor = factor(payor, levels = c("Private", "Public", "Other")),
    lvad_type = case_when(
      VadBrandId %in% c('236', '330') ~ "HeartMate 3",
      VadBrandId %in% c('205', '313') ~ "HeartMate II",
      VadBrandId %in% c('224', '316') ~ "HVAD",
      TRUE ~ "Other"),
    lvad_type = factor(lvad_type, levels = c("HeartMate 3", "HeartMate II", "HVAD", "Other"))
  )

cat("SRTR Table 1 data prepared:", nrow(srtr_table1), "patients\n")

# ============================================================================
# PREPARE STAR TABLE 1 DATA
# ============================================================================

star_table1 <- star_lvad_outcomes %>%
  select(WL_ID_CODE, first_lvad_implant) %>%
  rename(lvad_implant_date = first_lvad_implant) %>%
  left_join(
    thoracic_data_2025q2 %>%
      select(WL_ID_CODE, GENDER, ETHCAT, INIT_AGE, INIT_WGT_KG_CALC, 
             INIT_HGT_CM_CALC, ABO, TCR_DGN, PRI_PAYMENT_TCR, INIT_DATE),
    by = "WL_ID_CODE"
  ) %>%
  left_join(
    thoracic_mcs_device_2025q2 %>%
      filter(DEVICE_TYPE == "Dischargeable VAD") %>%
      filter(VENTRICULAR_SUPPORT == "Left") %>%
      group_by(WL_ID_CODE) %>%
      arrange(IMPLANT_DATE) %>%
      slice(1) %>%
      ungroup() %>%
      select(WL_ID_CODE, LVAD_BRAND, VAD_BRAND),
    by = "WL_ID_CODE"
  ) %>%
  mutate(
    data_source = "OPTN STAR",
    age_at_listing = floor(INIT_AGE),
    age_at_lvad = INIT_AGE + as.numeric(difftime(lvad_implant_date, INIT_DATE, units = "days")) / 365.25,
    age_at_lvad = floor(age_at_lvad),
    sex = ifelse(GENDER == "M", "Male", "Female"),
    sex = factor(sex, levels = c("Male", "Female")),
    race = case_when(
      ETHCAT == 1 ~ "White",
      ETHCAT == 2 ~ "Black",
      ETHCAT == 4 ~ "Hispanic/Latino",
      ETHCAT == 5 ~ "Asian",
      TRUE ~ "Other"),
    race = factor(race, levels = c("White", "Black", "Hispanic/Latino", "Asian", "Other")),
    bmi = INIT_WGT_KG_CALC / ((INIT_HGT_CM_CALC / 100) * (INIT_HGT_CM_CALC / 100)),
    bmi_categories = case_when(
      bmi < 18 ~ "Underweight",
      bmi >= 18 & bmi < 25 ~ "Normal",
      bmi >= 25 & bmi < 30 ~ "Overweight",
      bmi >= 30 ~ "Obese",
      TRUE ~ "Unknown"),
    bmi_categories = factor(bmi_categories, levels = c("Normal", "Underweight", "Overweight", "Obese", "Unknown")),
    blood_type = factor(
      case_when(
        ABO %in% c("A", "A1", "A2") ~ "A",
        ABO %in% c("A1B", "A2B") ~ "AB",
        TRUE ~ ABO)),
    diagnosis = case_when(
      TCR_DGN > 999 & TCR_DGN < 1007 ~ "Dilated Cardiomyopathy, Non-Ischemic",
      TCR_DGN == 1007 | TCR_DGN == 1200 ~ "Ischemic Cardiomyopathy",
      TCR_DGN > 1048 & TCR_DGN < 1100 ~ "Restrictive Cardiomyopathy",
      TCR_DGN == 1203 | (TCR_DGN >= 1205 & TCR_DGN <= 1207) ~ "Congenital Heart Disease",
      TRUE ~ "Other"),
    diagnosis = factor(diagnosis, levels = c("Dilated Cardiomyopathy, Non-Ischemic", "Ischemic Cardiomyopathy", 
                                             "Restrictive Cardiomyopathy", "Congenital Heart Disease", "Other")),
    payor = case_when(
      PRI_PAYMENT_TCR == 1 ~ "Private",
      PRI_PAYMENT_TCR %in% c(2, 3, 4, 5, 6, 7, 13) ~ "Public",
      TRUE ~ "Other"),
    payor = factor(payor, levels = c("Private", "Public", "Other")),
    device_brand = coalesce(VAD_BRAND, LVAD_BRAND),
    lvad_type = case_when(
      grepl("HeartMate 3|HM3|HM 3|HeartMate III", device_brand, ignore.case = TRUE) ~ "HeartMate 3",
      grepl("HeartMate II|HMII|HM II|HM2|HM 2", device_brand, ignore.case = TRUE) ~ "HeartMate II",
      grepl("HVAD|HeartWare", device_brand, ignore.case = TRUE) ~ "HVAD",
      TRUE ~ "Other"),
    lvad_type = factor(lvad_type, levels = c("HeartMate 3", "HeartMate II", "HVAD", "Other"))
  )

cat("STAR Table 1 data prepared:", nrow(star_table1), "patients\n")

# ============================================================================
# COMBINE AND CREATE TABLE 1 (NO P-VALUES)
# ============================================================================

combined_table1 <- bind_rows(
  srtr_table1 %>% select(data_source, age_at_listing, age_at_lvad, sex, race, bmi_categories, 
                         blood_type, diagnosis, payor, lvad_type),
  star_table1 %>% select(data_source, age_at_listing, age_at_lvad, sex, race, bmi_categories, 
                         blood_type, diagnosis, payor, lvad_type)
) %>%
  mutate(data_source = factor(data_source, levels = c("SRTR", "OPTN STAR")))

# Set variable labels
var_label_list <- list(
  age_at_listing = "Age at Listing (Years)",
  age_at_lvad = "Age at LVAD Implant (Years)",
  sex = "Sex",
  race = "Race/Ethnicity",
  bmi_categories = "BMI Category",
  blood_type = "Blood Type",
  diagnosis = "Primary Diagnosis",
  payor = "Insurance Type",
  lvad_type = "LVAD Device Type"
)

labelled::var_label(combined_table1) <- var_label_list

# Create table WITHOUT p-values
table1 <- combined_table1 %>%
  tbl_summary(
    by = data_source,
    digits = list(all_categorical() ~ c(0, 1)),
    missing = "no"
  ) %>%
  add_overall() %>%
  modify_header(label ~ "**Characteristic**") %>%
  modify_caption("**Table 1. Baseline Characteristics of LVAD Cohorts by Data Source**") %>%
  modify_footnote(all_stat_cols() ~ "Cohorts are not independent; statistical testing not performed") %>%
  bold_labels()

print(table1)

colSums(is.na(combined_table1))

```

``` {r redo flow anlaysis}

# ============================================================================
# STAR JUSTIFICATION FLOW ANALYSIS - COMPLETE CODE
# ============================================================================

# Flow function for STAR justification data
get_star_just_lvad_flow <- function(start_date, end_date) {
  candidates_in_period <- thoracic_data_2025q2 %>%
    filter(INIT_DATE <= end_date) %>%
    filter(is.na(END_DATE) | END_DATE > start_date) %>%
    filter(INIT_AGE > 17) %>%
    dplyr::select(WL_ID_CODE, INIT_DATE, END_DATE)
  
  candidates_with_lvad <- candidates_in_period %>%
    left_join(star_lvads_just %>% dplyr::select(WL_ID_CODE, first_lvad_implant), 
              by = "WL_ID_CODE")
  
  new_listings_with_lvad <- candidates_with_lvad %>%
    filter(INIT_DATE >= start_date & INIT_DATE <= end_date) %>%
    filter(!is.na(first_lvad_implant) & first_lvad_implant <= INIT_DATE) %>%
    nrow()
  
  lvad_implants_on_waitlist <- candidates_with_lvad %>%
    filter(!is.na(first_lvad_implant)) %>%
    filter(first_lvad_implant >= start_date & first_lvad_implant <= end_date) %>%
    filter(first_lvad_implant > INIT_DATE) %>%
    nrow()
  
  total_new_listings <- candidates_with_lvad %>%
    filter(INIT_DATE >= start_date & INIT_DATE <= end_date) %>%
    nrow()
  
  return(data.frame(
    period_start = start_date,
    period_end = end_date,
    new_listings_with_lvad = new_listings_with_lvad,
    lvad_implants_on_waitlist = lvad_implants_on_waitlist,
    total_new_listings = total_new_listings
  ))
}

# Define periods for STAR justification (Jan 1, 2019 - Apr 30, 2025)
period_starts_just <- seq(mdy("01-01-2019"), mdy("03-01-2025"), by = "2 months")
period_ends_just <- period_starts_just %m+% months(2) - days(1)
period_ends_just[length(period_ends_just)] <- mdy("04-30-2025")

# Run flow analysis
star_just_flow_list <- lapply(1:length(period_starts_just), function(i) {
  get_star_just_lvad_flow(period_starts_just[i], period_ends_just[i])
})

star_just_flow_df <- bind_rows(star_just_flow_list) %>%
  mutate(
    period_label = paste(format(period_start, "%b"), format(period_end, "%b %Y"), sep = "-"),
    period_factor = factor(period_label, levels = period_label),
    pct_listed_with_lvad = (new_listings_with_lvad / total_new_listings) * 100,
    total_lvad_incidence = new_listings_with_lvad + lvad_implants_on_waitlist
  )

# ============================================================================
# FILTER SRTR AND STAR MCS TO MATCH TIMEFRAME
# ============================================================================

# Add total_lvad_incidence if not already present
if (!"total_lvad_incidence" %in% colnames(flow_results_df)) {
  flow_results_df <- flow_results_df %>%
    mutate(total_lvad_incidence = new_listings_with_lvad + lvad_implants_on_waitlist)
}

if (!"total_lvad_incidence" %in% colnames(star_flow_df)) {
  star_flow_df <- star_flow_df %>%
    mutate(total_lvad_incidence = new_listings_with_lvad + lvad_implants_on_waitlist)
}

# Filter to match justification timeframe (Jan 2019 onwards)
flow_results_df_filtered <- flow_results_df %>%
  filter(period_start >= mdy("01-01-2019"))

star_flow_df_filtered <- star_flow_df %>%
  filter(period_start >= mdy("01-01-2019"))

# ============================================================================
# COMBINE ALL THREE SOURCES
# ============================================================================

combined_flow <- bind_rows(
  flow_results_df_filtered %>%
    dplyr::select(period_start, period_end, period_label, total_lvad_incidence, 
                  new_listings_with_lvad, lvad_implants_on_waitlist) %>%
    mutate(data_source = "SRTR (MCS File)"),
  star_just_flow_df %>%
    dplyr::select(period_start, period_end, period_label, total_lvad_incidence,
                  new_listings_with_lvad, lvad_implants_on_waitlist) %>%
    mutate(data_source = "STAR (MCS + Status Just. Files)"),
  star_flow_df_filtered %>%
    dplyr::select(period_start, period_end, period_label, total_lvad_incidence,
                  new_listings_with_lvad, lvad_implants_on_waitlist) %>%
    mutate(data_source = "STAR (MCS File)")
) %>%
  mutate(
    period_factor = factor(period_label, levels = unique(period_label)),
    data_source = factor(data_source, 
                         levels = c("SRTR (MCS File)", "STAR (MCS + Status Just. Files)", "STAR (MCS File)"))
  )

# ============================================================================
# THREE-WAY TOTAL INCIDENCE PLOT
# ============================================================================

flow_3way_plot <- ggplot(combined_flow,
                         aes(x = period_factor, y = total_lvad_incidence,
                             color = data_source, linetype = data_source,
                             group = data_source)) +
  geom_line(linewidth = 1.2) +
  geom_point(aes(shape = data_source), size = 2.5) +
  scale_color_manual(values = c("SRTR (MCS File)" = "coral",
                                "STAR (MCS + Status Just. Files)" = "darkgreen",
                                "STAR (MCS File)" = "steelblue")) +
  scale_linetype_manual(values = c("SRTR (MCS File)" = "solid",
                                   "STAR (MCS + Status Just. Files)" = "solid",
                                   "STAR (MCS File)" = "dashed")) +
  scale_shape_manual(values = c("SRTR (MCS File)" = 16,
                                "STAR (MCS + Status Just. Files)" = 15,
                                "STAR (MCS File)" = 17)) +
  scale_y_continuous(labels = comma, limits = c(0, NA)) +
  labs(
    x = "Time Period",
    y = "New LVAD Candidates",
    color = "Data Source",
    linetype = "Data Source",
    shape = "Data Source"
  ) +
  theme_classic(base_size = 15) +
  theme(
    legend.position = "bottom",
    axis.text.x = element_text(angle = 45, hjust = 1, size = 12),
    axis.text.y = element_text(size = 15),
    axis.title = element_text(size = 15),
    legend.text = element_text(size = 13)
  )

print(flow_3way_plot)

# ============================================================================
# FLOW COMPONENTS PLOT - ALL THREE DATA SOURCES
# ============================================================================

# Prepare breakdown data
flow_breakdown <- combined_flow %>%
  tidyr::pivot_longer(
    cols = c(new_listings_with_lvad, lvad_implants_on_waitlist),
    names_to = "component",
    values_to = "count"
  ) %>%
  mutate(
    component = factor(component,
                       levels = c("new_listings_with_lvad", "lvad_implants_on_waitlist"),
                       labels = c("Listed with LVAD", "LVAD Implanted While Listed"))
  )

# Create the plot
flow_components_3way <- ggplot(flow_breakdown,
                               aes(x = period_factor, y = count,
                                   color = data_source,
                                   linetype = component,
                                   group = interaction(data_source, component))) +
  geom_line(linewidth = 1.2) +
  geom_point(aes(shape = component), size = 2.5) +
  scale_color_manual(
    values = c(
      "SRTR (MCS File)" = "coral",
      "STAR (MCS + Status Just. Files)" = "darkgreen",
      "STAR (MCS File)" = "steelblue"
    )
  ) +
  scale_linetype_manual(
    values = c("Listed with LVAD" = "solid",
               "LVAD Implanted While Listed" = "dotted")
  ) +
  scale_shape_manual(
    values = c("Listed with LVAD" = 16,
               "LVAD Implanted While Listed" = 17)
  ) +
  scale_y_continuous(labels = comma, limits = c(0, NA)) +
  labs(
    x = "Time Period",
    y = "Number of Candidates",
    color = "Data Source",
    linetype = "Event Type",
    shape = "Event Type"
  ) +
  theme_classic(base_size = 15) +
  theme(
    legend.position = "bottom",
    legend.box = "vertical",  # Stack legends vertically
    axis.text.x = element_text(angle = 45, hjust = 1, size = 15),
    axis.text.y = element_text(size = 15),
    axis.title = element_text(size = 15),
    legend.text = element_text(size = 15)
  ) +
  guides(
    color = guide_legend(order = 1, title = "Data Source"),
    linetype = guide_legend(order = 2, title = "Event Type"),
    shape = guide_legend(order = 2, title = "Event Type")
  )
print(flow_components_3way)


```
